---
title: "BDA project 2025-2026"
author: "J.A. Westerhuis, A.U.S. Heintz-Buschart"
date: "06/01/2026 - 17/01/2026"
output:
  pdf_document:
    number_sections: no
    toc: yes
    toc_depth: 3
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
editor_options:
  markdown:
    wrap: 72
bibliography: references.bib
---

\newpage

# BDA project 2025-2026

## Changes to the first part

The following questions were revised after the initial feedback:

-   **Question 1F** (CHANGED)
-   **Question 2B** (CHANGED)
-   **Question 2D** (CHANGED)
-   **Question 2K** (CHANGED)
-   **Question 3D** (CHANGED)

## Project description

Welcome to the BDA project 2025-2026 syllabus. In this document you will
find all the information you need to do your BDA project.\

The main goals of the Project are to learn how to:

1.  analyze your data using common data analysis methods,\
2.  visualize the results of the analysis methods,\
3.  be able to interpret the results.\

In this course you can use Quarto or R-Markdown to combine explainable
text and code in the same document. You have to submit a PDF by
rendering of knitting your document. Make sure your PDF only contains
the necessary text and code and not e.g. a print of all your data.

Check the Statistics with R syllabus for tips on working reproducible!

## Exercise description

Besides this project, every topic also has exercises. The goal of the
exercises is to teach you the different properties of the data and of
the data analysis methods:

1.  What are the properties of each sample and how do samples relate to
    each other ?\

2.  What are the properties of the variables, and how are they related
    ?\

3.  What is the distribution of the samples for certain variables ?\

4.  What is the noise level, how much systematic variation is present?\

5.  What is the meaning of scores, loadings, residuals in PCA?\

6.  Which distance measures should be used in clustering methods.\

7.  How to validate the result of a classification method.\

Thus the goal of the exercise is to make you aware of the specifics of
the data and the methods, while the Project is to show you are able to
use R to apply the methods, visualize and interpret results. Therefore,
it is important to first go through the exercises, discuss with your
fellow students, get feedback from TAs and teachers such that you
understand the data and the methods.

NB!

1.  You are responsible for sufficient interaction and feedback.\
2.  In answering the Project questions plagiarism rules still apply,
    thus if you use information other than your own ideas, give proper
    reference.\

### Groups

The Project has to be done in pairs. This is because many of the
open-ended questions we ask during your project are open for debate. In
such cases it is really helpful to have a partner who you can discuss
with about what the next step should be. Also you will be able to run
RStudio on two computers at the same time, which will help you in
splitting up the tasks and completing the assignments faster overall.
Please supply your names in the code block below.

```{r Names, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
name1 = "Thomas Uijen"
name2 = "Sara Lopez"
```

### Submission and grading

You will hand in your project quarto or markdown file and knitted .pdf
files in the first and second week. In the first week you will get an
indicative pass/fail grade and some feedback to help you progress. In
the second week you will receive a full grade and some feedback. The
deadline for the first week submission is Tuesday January 13 8:00 (AM),
for which you need to complete Assignments 1 up to and including
Assignment 3. The deadline for the second week submission is Tuesday
January 20 08:00 (AM), for which you will need to complete Assignment 1
up to and including Assignment 6. Note that you can use the week 1
feedback to improve your answers from Assignments 1-3 before the second
submission! If you do, then you have to indicate in the new PDF what was
changed from the previous submission. The latest version of Assignments
1-3 will be used for final grading in week 2. Please refer to the course
book on Canvas for more information on the BDA course organisation.

Grading of the project will be done using a rubric for each assignment.
An overview of the number of points you can get for every assignment is
given below. Every question states how many points you can get for it.

| Assignment                      | Number of points |
|---------------------------------|------------------|
| 1: Linear Algebra               | 20               |
| 2: Gene Expression Analysis     | 20               |
| 3: Principal Component Analysis | 20               |
| 4: Clustering                   | 20               |
| 5: Classification               | 20               |
| 6: ASCA                         | 20               |
| Total                           | 120              |

It is crucial for the graders to understand what you have done and why.
Please elaborate in the supplied text boxes below each question what
your reasoning is for making a certain step. To make your figures
eligible for grading, you should add a clear labels and titles, as well
as a clear description of what you have done to make your figure. When
writing your code in the code blocks, add comments clarifying what you
are doing!

### Root dir

To make sure R knows where to find the data and where to store results
you have to set a root directory. Fill in here the directory on your
computer where the data is stored.

```{r Rootdir, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
knitr::opts_knit$set(root.dir = '~/Documents/MSc_Sys_Bio/BDA/Project')
```

The following code block contains some extra settings to make sure that
your deliverable at the end of the project will be easier to grade for
the teaching team. Please do not modify it in any way.

```{r truncate output, echo=FALSE, include=FALSE, message=FALSE}
# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
      # truncate the output
      x <- c(head(x, n/2), '....', tail(x, n/2 + 1), '\n')
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
```

### Help

As any programmer will tell you, looking stuff up on Google is a major
part of any project. Hence looking up specific R programming questions
may be helpful in progressing in your project assignments. Coding
websites such as [StackOverflow](https://stackoverflow.com/) are
particularly helpful.

Specific questions on how functions work in R can be answered using the
R documentation. Say you want to understand how you can run the singular
value decomposition function svd(). On the right side of your screen you
can find a "help" tab where you can search for the function that you
want to use. You can type "svd" in the search field there to reach the
documentation of the svd() function. Additionally, you can type "?svd"
in your console window at the bottom of your screen to automatically
search for the documentation of your supplied function.

If things are unclear to you or you have any other question, feel free
to ask any of your lecturers or teaching assistants (TAs):\
- Johan Westerhuis (Course coordinator, teacher
[j.a.westerhuis\@uva.nl](mailto:j.a.westerhuis@uva.nl){.email})\
- Anna Heintz Buschart (Teacher,
[a.u.s.heintzbuschart\@uva.nl](mailto:a.u.s.heintzbuschart@uva.nl){.email})\
- Gustavo Stolf Jeuken (Teacher,
[g.stolfjeuken\@vu.nl](mailto:g.stolfjeuken@vu.nl){.email})\
- Alex Sanchez Cano (Teacher,
[a.sanchezcano\@uva.nl](mailto:a.sanchezcano@uva.nl){.email})\

### Installation of necessary packages

Below are the packages we will use throughout the course. Do not edit
this code block!

```{r load packages, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE}

if(!"umap" %in% installed.packages()[,1]){
  install.packages("umap")
}

if(!"MASS" %in% installed.packages()[,1]){
  install.packages("MASS")
}

if(!"gtools" %in% installed.packages()[,1]){
  install.packages("gtools")
}

if(!"stringr" %in% installed.packages()[,1]){
  install.packages("stringr")
}

library(umap)
library(MASS)
library(gtools)
library(stringr)
```

### The MoTrPAC data

The Molecular Transducers of Physical Activity Consortium (MoTrPAC)
profiled the dynamic change of multiple omics layers in several organs
of rats undergoing physical training. The results are presented in the
Nature article "Temporal dynamics of the multi-omic response to
endurance exercise training". The paper can be found on the canvas page,
and the data was obtained from their website MoTrPAC-data.org.

### Experimental design

Six-month-old male and female Fischer 344 rats were subjected to
progressive treadmill endurance exercise training for 1, 2, 4 or 8
weeks. Sex-matched sedentary, untrained rats were used as controls. In
this project we will focus on the transcriptomics data obtained from
blood cells and metabolomics data from plasma. Unfortunately these
measurements do not all come from the same rats.

## Assignments

In this assignment, you will start your exploration of the data by
checking various summaries per row and per column of your data tables.
You will also make a few histograms and inspect them to check the
comparability of the samples and variables. Finally you will mean center
and scale your data for use in the other assignments. NB! We assume you
have worked through the R-tutorial before starting this assignment.

### First look at the data

Load the metabolomics and transcriptomics data sets.

```{r Data importing, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_metabolomics = read.csv("Plasma_Metabolomics_Normalized.csv", row.names = 1)
df_transcriptomics = read.csv("Blood_Transcriptomics_Counts.csv",row.names = 1)
```

Take a quick view of the different data sets. Lets start with the
metabolomics data.

```{r metabolomics data example, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
print(df_metabolomics[1:7, c(1:8)])
```

In the first few columns are numbers identifying the specific rat
number, the experimental group (training or control) and experimental
time (1,2,4,8 weeks) as well as the sex of the rat (1 = female, 2 =
male). From column 4 onwards is where the metabolomics data is shown.
For the metabolomics data, the metabolite levels are normalized by the
measurement of a known standard.

Then take a quick view of the transcriptomics data.

```{r transcriptomics data example, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
print(df_transcriptomics[1:7, c(1:8)])
```

The transcriptomics data is transposed, the rats are in the columns
while the genes are in the rows. The experimental details for each rat
can be found in the last few rows of the data frame. Furthermore, the
counts are not considered numeric. Let's change that.

```{r Transpose Transcriptomics data}
df_transcriptomics = data.frame(t(df_transcriptomics))
print(df_transcriptomics[1:7, c(1:8)])
```

### From now on you have to give the code and answers yourself!

### Assignment 1: Linear Algebra (Johan Westerhuis, 05-01-2026)

#### 1A) Data pretreatment:

1)  For the metabolomics data, remove the columns with experimental
    details and Sex of the rat and store them in a new data frame
    df_metabolomics_Exp. Make a new column in the experimental data
    frame called Groups indicating one of these 5 levels: Control 8
    weeks, Training 1 week, Training 2 weeks, Training 4 weeks and
    Training 8 weeks.[1]

2)  Also make a new data frame df_metabolomics_Values with only the
    metabolite levels.[1]

3)  Remove metabolite Oleoyl.EA as this one has missing values for some
    rats. [1]

    !N.B. Do not save the changed file
    "Plasma_Metabolomics_Normalized.csv". At the end of this assignment
    we will save the data under a different name such that
    "Plasma_Metabolomics_Normalized.csv" will not be changed.

```{r Assignment 1A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggrepel)
## 1
df_metabolomics_Exp<-df_metabolomics|>
  dplyr::select(1:3)|>
  unite(col = "Groups",Experiment.Group,Experiment.Time,sep=" ",remove = F)


## 2/3
df_metabolomics_Values<-df_metabolomics|>
  dplyr::select(-c(1:3))|>
  dplyr::select(-Oleoyl.EA)



```

#### 1B)

For the transcriptomics also make data frames with only experimental
details and only with values. Note the experimental details can be found
at the end of the data frame. Also make and add the Group variable to
the df with experimental details [1].

```{r Assignment 1B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_transcriptomics_Exp<-df_transcriptomics|>
  dplyr::select(Experiment.Group,Experiment.Time,sex)|>
  unite(Experiment.Group,Experiment.Time,col="Groups",sep=" ",remove =F)

df_transcriptomics_Values<-df_transcriptomics|>
  dplyr::select(-c(Experiment.Group,Experiment.Time,sex))
```

#### To make sure the transcriptomics values are considered numeric, run the next line:

```{r Make transcriptomics counts numeric}
df_transcriptomics_Values = as.data.frame(apply(df_transcriptomics_Values, 2, function(x) as.numeric(as.character(x))))
```

#### 1C)

Many genes only have counts in few rats. Find genes that have nonzero
counts for more than 25 rats. Use the apply function to first calculate
for each gene the number of samples having nonzero counts. Then select
those genes for which this number is larger than 25. Make a new data
frame df_transcriptomics_Filtered that only contains these genes. [1]

```{r Assignment 1C,Filter genes, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# Compute number of non-zero counts per gene
non_zero_genes<-apply(df_transcriptomics_Values, 2, function(x){sum(x!=0)})

# Filter transcriptomics data
df_transcriptomics_Filtered<-df_transcriptomics_Values[,non_zero_genes>25]

# Verify procedure
!any(apply(df_transcriptomics_Filtered, 2, function(x){sum(x!=0)})<=25)

```

#### 1D)

Now write the metabolomics_Values, metabolomics_Exp,
transcriptomics_Filtered and transcriptomics_Exp data frames into new
csv files in your working directory. [2]

```{r Assignment 1D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
dfs<-list(df_metabolomics_Exp=df_metabolomics_Exp,
          df_metabolomics_Values=df_metabolomics_Values,
          df_transcriptomics_Filtered=df_transcriptomics_Filtered,
          df_transcriptomics_Exp=df_transcriptomics_Exp)

lapply(names(dfs), function(nm) {
  write.csv(dfs[[nm]], paste0(nm, ".csv"), row.names = TRUE)
})
```

Check whether the files were written correctly. Now you have cleaned
metabolomics and transcriptomics data which you will explore in the next
assignments.

#### 1E)

1.  Calculate the total sum of the normalized values per row and per
    column of the filtered transcriptome data set using the rowSums()
    and colSums() functions. [1]\
2.  Plot the log of the row and column sums of the filtered
    transcriptomics data in a histogram in a figure with 2 subfigures.
    Add labels for the x-axis and titles for each subfigure. [1]\

```{r Assignment 1E, include=TRUE, eval=TRUE, echo=TRUE}
## 1

sum_per_rat<-rowSums(df_transcriptomics_Filtered)
sum_per_gene<-colSums(df_transcriptomics_Filtered)


## 2
df<-data.frame(value=c(sum_per_rat,sum_per_gene),
           group=c(rep("Log-sum per rat",length(sum_per_rat)),rep("Log-sum per gene",length(sum_per_gene))))

ggplot(df,aes(x=log(value),after_stat(density)))+geom_histogram(fill="grey",colour="black")+facet_wrap(~group,scales="free")+labs(x="Log-sum",title="Histograms for the log sum per gene and per rat")



```

#### 1F): Interpretation. (Fixed)

Write down for each plot what you can learn about the samples and
variables in the data sets.[2]\
Add information of - the meaning of a row sum and of a column sum for
these data. - what does it mean when the sum of all values of a sample
or of a feature is larger than of another sample or feature? - Why is it
useful to plot the log of the row and column sums?

#### Put any text for Interpretation 1F here (Fixed). *CHANGED*

1)  In the first histogram we learn about the samples: The rowSum is the
    total expression per sample, hence the overall signal across all
    genes. We observe a narrow distribution in the dataset, suggesting
    that most of the samples have similar total expression, meaning that
    the measurements are consistent across samples. Values lying in the
    tails may suggest potentials outliers, this means unusually low or
    high total expression (overall signal) which mainly reflect the
    biological effects on the gene expressions. While some random noise
    is still present, by log-transforming the data we compensate for
    systematic error and bias.

In the second histogram we learn about variables: The colSum represents
the total expression per gene across samples. We observe a wide
distribution of log-transformed colSums, indicating strong heterogeneity
among genes, hence some genes are highly expressed across the samples
while others are lowly expressed. This depicts that genes contribute
differently to the overall transcriptional signal, this is a common
feature of transcriptomics data sets. The plot also shows a long right
tail, indicating a small subset of very highly expressed genes, this is
normally expected in transcriptomics since most genes are expressed at
low levels, while only a small number of housekeeping genes are
expressed at very high levels, hence creating a right-skewed
distribution.

2)  When the sum of all values of a sample is larger than that of
    another sample that suggests a higher total measured expression in
    that precise sample, hence higher sequencing depth (or: overall
    transcriptional activity). In the case that the sum of all values of
    a feature in our case a gene is larger than that of another feature,
    it means that this gene is more abundantly expressed across the
    dataset, probably because it is expressed at higher levels or
    because it is expressed in more samples.

3)  It is useful because the sums either for columns or rows vary over a
    large/wide range of values and the log transformation reduces the
    impact of extreme values, making the distributions easier not only
    to visualize but also to compare, this allows patterns and potential
    outliers to be identified more clearly.

#### End of text block.

#### 1G)

1.  Calculate the column mean and column standard deviation for the
    filtered transcriptomics data sets using the apply() function. [1]\
2.  Make a scatter plot in which the log(standard deviation) is plotted
    vs the log(mean) of each variable for the transcriptomics
    dataset.[1]\

```{r Assignment 1G, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
## 1
col_means<-apply(df_transcriptomics_Filtered,2,mean)
col_sds<-apply(df_transcriptomics_Filtered,2,sd)

df<-tibble(col_means=col_means,col_sds<-col_sds)

ggplot(df,aes(x=log(col_means),y=log(col_sds)))+geom_point(size=0.8)+labs(x="log(Mean)",y="log(Standard deviation)",title = "Mean-variance relationship of gene expression (log scale)")+theme(plot.title = element_text(hjust = 0.5))
```

#### 1H): Interpretation.

1.  Explain what is the meaning of the standard deviation of a gene.
    Take into account the knowledge you have about the samples [1].
2.  Explain what information about the data is provided by the
    scatterplot of Exercise 1G. [1]\

#### Put your interpretation 1H here.

1)  The standard deviation of a gene measures how much of the expression
    level of a gene "x" varies across the samples. A low standard
    deviation indicates that the expression levels of that gene is
    similar across all samples (rats), suggesting stable expression.
    Whereas a high standard deviation indicates a strong variability in
    gene expression between samples, this could reflect biological
    differences between conditions , in this case :sex or training
    period or differential gene regulation across rats.

Since the dataset contains samples from different conditions, the
standard deviation reflect mainly biological differences in gene
expression between samples, rather than just technical variation/errors.

2)  The scatterplot shows the relationship between the mean expression
    and the standard deviation of each gene on a log scale. There is a
    clear positive trend which indicates that those genes with higher
    mean expression tend to have a higher variability across the
    samples.This is a common feature of transcriptomics dataset and
    indicates that variability in gene expression often increases with
    expression level.

Moreover, the plot also depicts that lowly expressed genes generally
have low variability, while the highly ones span a wider range of
standard deviations. This plot highlights that gene expression data are
heteroscedastic, meaning that variability differs between genes, this is
a pivotal aspect to take into account when applying statistical methods
that assume equal variance.

#### End of text block.

#### 1I)

Now we go back to the metabolomics data.

1\. Make a new variable that only contains the Alanine levels of the
rats. [0.5]\
2. Meancenter the Alanine levels of the new variable. [0.5]\
3. Scale the meancentered Alanine levels of the new variable with its
standard deviation. [0.5]\
4. Plot the raw, only meancentered and meancentered and scaled Alanine.
levels in a figure with 3 subplots and add the correct labels on the
axes. [0.5]\

#### Put your code here

```{r Assignment 1I, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
Alanine_levels<-df_metabolomics_Values|>
  select(Alanine)|> ## 1 is this the correct value, I see multiple?)
  mutate(Raw=Alanine,
         Centered=Alanine-mean(Alanine),
        `Centered and Scaled`=(Alanine-mean(Alanine))/sd(Alanine))|> ## 2/3
  select(-Alanine)|>
  pivot_longer(cols=3:1,values_to = "values",names_to = "processing")|>
  mutate(
    processing = factor(
      processing,
      levels = c("Raw", "Centered", "Centered and Scaled")
    )
  )


## 4

ggplot(Alanine_levels,aes(values, after_stat(density)))+geom_histogram(bins = 11,fill="grey",colour="black")+facet_wrap(~processing,scales="free")+labs(x="Alanine expression level",title= "Histograms for Alanine expression with different transformations")+theme(plot.title = element_text(hjust = 0.5))



```

#### 1J): Interpretation.

1.  Explain what centering does to your variable. [1]\
2.  Explain what the combination of centering and scaling does to your
    data. [1]\

#### Put any text for interpretation 1J here.

1)  Centering substracts the mean of the variable from each sample,
    hence shifting all values so that the average becomes zero, but the
    differences between samples remain the same. This results in
    positive and negative values indicating whether a sample has higher
    or lower Alanine levels compared to the average.

2)  When combining centering and scaling we standardize the data by
    setting the mean to zero and the standard deviation to one. This
    removes differences in the average level while preserving the
    relative differences between samples and making the values
    comparable, because after scaling, differences between samples are
    expressed in units of standard deviation, which is useful when
    comparing variables measured on different scales or when applying
    multivariate statistical methods.

#### End of text block.

The combination of meancentering and scaling by the standard deviation
of each column in a matrix is often used in data analysis. It is known
by different names, e.g. unit variance scaling, auto scaling,
standardization. The standardized values are also known as z-scores.

#### 1K)

Standard deviations of metabolite levels.

1\. Calculate the standard deviation of each metabolite over all samples
and make a histogram of the standard deviations of the metabolomics
columns, [1].

2\. Discuss what you can learn from the histogram, and how this
information is used with respect to scaling the data. [1]

```{r Assignment 1K, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
## 1 
sd_metabolomics<-tibble("sd"=apply(df_metabolomics_Values,2,sd))
ggplot(sd_metabolomics,aes(sd,after_stat(density)))+geom_histogram(bins = 15,fill="grey",colour="black")+labs(x="Standard deviation of metabolites across samples",title="Distribution of metabolite standard deviation")+theme(plot.title = element_text(hjust = 0.5))
```

**Put your discussion what you can learn for the histogram here.**

The histogram shows that metabolites differ considerably in their
standard deviation across samples. Most metabolites have low
variability, while a smaller subset shows much higher variability,
resulting in a right-skewed distribution. This indicates a very
different levels of variability across samples with some metabolites
changing only slightly across samples whereas others show strong
variation.

This is one of the reasons why scaling is crucial, because by dividing
each metabolite by its standard deviation ensures that all metabolites
contribute equally to the analysis.

**End of discussion**

### This concludes Assignment 1.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session

### Assignment 2: Gene expression analysis (Douwe Molenaar, 6-01-2026)

In the previous assignment, you explored the expression data of the rats
in terms of counts of mRNA of specific genes. In this assignment we will
look for genes that are different between the Control group and the
Training group at 8 weeks. From the MotrPac website, the
Blood_RNA_Normalized_C8_T8.csv data file was obtained that only consists
of the rats of the training group at 8 weeks and the rats of the control
group. Specific normalization was already applied. In assignment 2 we
will use these data.

Load the normalized datafile Blood_RNA_Normalized_C8_T8.csv from Canvas.

```{r load normalized gene expression data }
df_transcriptomics_Filtered_Norm = read.csv("Blood_RNA_Normalized_C8_T8.csv", row.names = 1)
```

#### 2A)

Transpose the data and remove the sex variable.[1]

```{r Assignment 2A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
df_transcriptomics_Filtered_Norm<-data.frame(t(df_transcriptomics_Filtered_Norm))|>
  dplyr::select(-sex)
```

#### 2B) (fixed)

1.  Check whether the data is homoscedastic by plotting the log of the
    std vs the log of the mean per group.[1] *CHANGED*
2.  Make indexes for the control group and the 8week training group[0.5]
3.  Calculate the mean and std per group for each gene.[0.5]
4.  Plot the std vs the mean in a loglog plot.[1] *CHANGED*

```{r Assignment 2B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}


#2
# Make indices 
control_group=row.names(df_transcriptomics_Filtered_Norm)[1:10]
training_group=row.names(df_transcriptomics_Filtered_Norm)[11:20]

#3
sd_mean_per_gene<-df_transcriptomics_Filtered_Norm|>
  mutate(Group=row.names(df_transcriptomics_Filtered_Norm))|>
  mutate(Group = factor(ifelse(Group %in% control_group, "Control", "Training")))|>
  pivot_longer(col=-Group,names_to = "Gene",values_to = "Value")|>
  group_by(Gene,Group)|>
  summarise(mean=mean(Value),sd=sd(Value))
  
#4
# Plot the sd and mean with pseudolog transform due to the previous normalisation
ggplot(sd_mean_per_gene,aes(x=mean,y=sd,colour=Group))+facet_wrap(~Group)+
  geom_point(size=0.8,alpha=0.5) +
 labs(title = "Mean-Variance relationship for log gene expression per group",x="Mean",y="Standard deviation")+theme(plot.title = element_text(hjust = 0.5))

ggplot(sd_mean_per_gene,aes(x=mean,y=sd,colour=Group))+
  geom_point(size=0.8,alpha=0.5) +
 labs(title = "Mean-Variance relationship for log gene expression",x="Mean",y="Standard deviation")+theme(plot.title = element_text(hjust = 0.5))
```

#### 2C: Interpretation

1.  Explain what kind of figure you would have expected for
    homoscedastic data.[1]
2.  Conclude about the homoscedasticity of this data based on the figure
    you obtained.[1]

#### Give your explanation and conclusion here (fixed)

Interpretation 1) For homoscedastic data, the variance is constant, thus
it does not vary with the mean expression level. In a mean-variance plot
on a log-log scale (as shown in plot1 and 2) for control and training
groups we would expect the data points to form a horizontal band, with
the standard deviation being approximately constant across the entire
range of mean expression values. Consequently we would not expect to
observe a systematic trend between the mean and the variance.

2)  Both plots (control and treatment) do not represent the expected
    pattern for homoscedastic data. In both the control and training
    groups, a clear mean-variance dependency is observed, characterized
    by a steep positive relation for mean \< -1, followed by a negative
    trend in which the standard deviation decreases as the mean
    expression increases, leading to a heel-shaped pattern. This
    indicates that the variance is not constant across expression
    levels,and therefore the data are heteroscedastic.

When comparing both groups, the control group shows a marginally more
compact distribution of points, particularly at mid-range to high
expression levels, suggesting a more stable variance structure.
However,a clear dependence of variance on mean expression remains
present. The general mean-variance relationship is shared by both
groups, although minute differences in dispersion are observed.

In general, the data are not homoscedastic.Instead, they reveal a strong
mean-variance dependency, which is a typical feature of gene expression
data. This highlights the need for statistical methods that explicitly
account for heteroscedasticity in data analysis.

In the next assignments we will apply tests to find genes related to
training by comparing the control group with the group of rats that
trained 8 weeks.

#### 2D) [Unfixed as per discord message) *CHANGED*

First we filter Genes for which there is almost no difference between
the samples (standard deviation \< 0.1) as these will provide problems
in the t-test. This can occur when e.g. for one of the group all rats
had zero counts. 1. Calculate the standard deviation for each gene over
all samples.[1] 2. Only keep those genes with a standard deviation
larger than 0.1.[1]

```{r Assignment 2D , include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# Collect genes that have sd>0.1
sd_genes<-apply(df_transcriptomics_Filtered_Norm,2,'sd')

# Select these genes
selected_genes_df<-df_transcriptomics_Filtered_Norm[,sd_genes>0.1]
```

#### 2E)

Perform a t-test of the expression of Gene ENSRNOG00000000024 between
the Control and Training 8 weeks conditions. [1]\

```{r Assignment 2E , include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

# Select control and treatment groups for this gene
control<-selected_genes_df$ENSRNOG00000000024[1:10]
treatment<-selected_genes_df$ENSRNOG00000000024[11:20]

#perform ttest
t.test(control,treatment,var.equal=F)
```

#### 2F) Interpretation.

1.  What is the hypothesis being tested? [1]\
2.  What does this p-value mean? [0.5]\
3.  What is your conclusion from the test result? [0.5]\

#### Put any text for Interpretation 2F here.

1)  The t-test evaluates whether the mean expression level of the gene
    ENSRNOG00000000024 differs between the Control and Training groups.
    More precisely, we have the following hypotheses:

The null hypothesis (H0): The mean gene expression is equal in the
control and training groups Alternative hypothesis (H1): The mean gene
expression is different between control an training groups.

2)  The obtained p-value (p=0.01292) represents the probability of
    measuring the observed difference in mean gene expression, or a more
    extreme difference, under the assumption that the null hypothesis is
    true.

3)  Since the p-value is less than the conventional significance
    threshold of 0.05 we conclude that there is a mean difference
    between both groups, as this p-value indicates that such a
    difference is unlikely to have arisen by chance. However, when
    considering multiple tests, it is important to recall that the false
    discovery rate is a more robust measure of accuracy than a p-value.

Since the obtained p-value is below the commonly used significance
threshold of 0.05, the null hypothesis is rejected. Statistically, this
suggests that the mean expression of ENSRNOG00000000024 differs between
the control and training groups at 8 weeks.

Moreover the sample means show that the gene has a higher average
expression in the training group (mean=3.20) compared to the control
group (mean=2.66). Furthermore, the 95% confidence interval for the
difference in means (-0.94 to -0.12) does not include zero, reinforcing
the conclusions that the observed difference it is statistically
significant.

This difference is corroborated by mechanistic knowledge, as this gene
codes for the protein Hebp1; a heme binding protein ([Ensembl (Hebp1,
Rattus
norvegicus)](https://www.ensembl.org/Rattus_norvegicus/Gene/Summary?g=ENSRNOG00000000024;r=4:169705668-169735199;t=ENSRNOT00000000025 "Hebp1 gene in Rattus norvegicus (Ensembl)").
According to UniProt, this protein "may bind free porphyrinogens that
may be present in the cell and thus facilitate removal of these
potentially toxic compounds"([Hebp1
function](https://www.uniprot.org/uniprotkb/B4F7C7/entry#function "UniProt entry for Hebp1").)

Overall, the results indicate that the expression of ENSRNOG00000000024
is significantly altered by training, rather than the observed
difference being due to random variation alone.

#### End of text block.

#### 2G)

1.  Perform a t-test for every gene in the gene expression dataset for
    Control and 8 weeks training using the apply() function, and store
    the p-value output of the t-test in a vector. [0.5]\
2.  Which gene has the smallest p-value? [0.5]

```{r Assignment 2G, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

# Computing all p-values
p_values=apply(selected_genes_df,2,function (x) {
  t.test(x[1:10],x[11:20])$p.value
  }
  )

# Selecting the smallest p-value
sort(p_values)[1]
```

#### Put any text for Interpretation 2G here.

The smallest p-value obtained from the t-test is
$p\approx3\times 10^{-6}$, corresponding to the gene
"ENSRNOG00000005576". This extremely low p-value indicates very strong
statistical evidence against the null hypothesis, meaning that the
probability of observing such a difference of at least this size in gene
expression, under the assumption that the null hypothesis is true, is
extremely small. Therefore, this gene shows a highly significant
difference in expression between the control and training groups at 8
weeks.

Nevertheless,a large number of genes were tested simultaneously, meaning
that this result should be carefully interpreted . Since multiple
testing increases the likelihood of false positives, appropriate
multiple-testing correction methods, such as false discovery rate (FDR)
correction, should be applied to verify the reliability of this finding.

To this effect, we compare this statistical result to known literature
of the gene, which codes for the enzyme Ribose-5-phosphate isomerase.
This protein plays a role in cellular metabolism, through the pentose
phosphate pathway [@b2024].

#### 2H)

1.  Make a histogram of all the p-values. Make a rough guess of how many
    positives are present in your set of genes? [1]

<!-- -->

2)  How many false positives do you expect given the p-value histogram?
    [1]

```{r Assignment 2H, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
## make a dataframe of the p-values
p_value_df<-data.frame(
  Gene = names(p_values),
  p_value = p_values
)


# Number of bins that we will use to find the number of positives
nbins <- 40 

#roughly 2800 positives

# Setting the boundaries of the bins
bin.breaks = seq(0, 1, by = 1/nbins)

#creating the plot
p<-ggplot(p_value_df,aes(p_value))+geom_histogram(breaks=bin.breaks)+theme(plot.title = element_text(hjust = 0.5))

# adding the approximate null level of negatives
null_level<-sum(p_value_df$p_value>=0.5)/sum(bin.breaks>0.5)

p+geom_hline(yintercept =  null_level,colour="red")+geom_vline(xintercept=0.05,colour="green")+theme(plot.title = element_text(hjust = 0.5))+labs(x="p-value",title="Histogram of the gene-wise p-values")

number_fp=null_level*nbins*0.05
number_fp

round(100*number_fp/sum(p_value_df$p_value<0.05),1)
```

#### Put any text for interpretation 2H here.

1)  From the plot we see that there are roughly 2750 positives: the
    first two bins (left of the green line marking the 0.05 p-value
    threshold) contain \~2000 and \~750 respectively.

2)  To compute the approximate number of false positives, we estimated
    the amount of negatives per bin by averaging the tail with p-value
    0.5 and up. This null level (shown in the plot in red) was then
    multiplied by 2, the number of bins with a p-value less than 0.05.
    This yields an estimated number of false positives of 481, an FDR of
    around 17,3%. This section of tests is represented in the graph by
    the area below the red line and left of the green line.

#### End of text block.

Our gene expression data is already normalized and log transformed, so
we can instead calculate the fold change using the averages between the
two conditions.\

#### 2I)

Calculate the fold change of all genes between Control and 8 weeks
training conditions this way. [1]

```{r Assignment 2I, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
control_mean<-sd_mean_per_gene|>
  filter(Group=="Control")|>
  dplyr::select(Gene,mean)|>
  filter(Gene %in% colnames(selected_genes_df))
  


training_mean<-sd_mean_per_gene|>
  filter(Group=="Training")|>
  dplyr::select(Gene,mean)|>
  filter(Gene %in% colnames(selected_genes_df))

FC<-data.frame("Gene"=colnames(selected_genes_df),"FC"=training_mean$mean-control_mean$mean)
```

As you may be aware, you have now done almost 14.000 t-tests. As such,
you will need to control for Type I error. In the exercises you did
before the project part of today's computer practical you have
calculated the FDR value for any p-value and plotted them. An easy
function to obtain the FDR values is p.adjust().

#### 2J)

1.  Calculate the adjusted p-values using the p-adjust() function. [1]\
2.  Make a volcano plot using your Fold change and adjusted p-values for
    every gene in a similar way as was done in the exercises of this
    topic. [1]\
3.  Set a threshold for both the adjusted p-value and for the Fold
    Change, add lines to the plot indicating the threshold values and
    show the genes that are differentially expressed according to your
    thresholds in a different color. [1]\

```{r Assignment 2J, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# Calculate adjusted p-values
p_value_adjusted_df<-p_value_df|>
  mutate(p_adjusted=p.adjust(p_value,"fdr"))|>
  left_join(FC)|>
  mutate(Diff_Exp=ifelse(abs(FC)>1& p_adjusted<0.05,"Differential","Identical"))
# Make Volcano plot
plot<-ggplot(p_value_adjusted_df,aes(x=FC,y= 1/p_adjusted ))+
  labs(x="Log fold-change of gene expression",y="(adjusted p-value)^-1",title = "Volcano plot of differential gene expression",colour="Gene expression")+theme(plot.title = element_text(hjust = 0.5))
points<-geom_point(size=0.5,alpha=0.5)
plot+points
# set thresholds and find differentially expressed genes
p_threshold<-0.05
FC_threshold=1

plot+geom_point(size=0.8,alpha=0.6,aes(colour=Diff_Exp))+
  geom_hline(yintercept = 1/p_threshold, col='red', linetype="dashed") +
  geom_vline(xintercept = c(-FC_threshold,FC_threshold), col='gray') +
  annotate("text", x=2.4, y=1/0.044, label="p-value < 0.05") +
  annotate("text", x=c(-1.5,1.5), y=c(51,51), label=c("log(FC) < -1", "log(FC) > 1"))+
  scale_color_manual(
    values = c(
      "Differential" = "red",
      "Identical" = "black"
    )
  )+
  labs(x="Log fold-change of gene expression",y="(adjusted p-value)^-1",title = "Volcano plot of differential gene expression",colour="Gene expression")+theme(plot.title = element_text(hjust = 0.5))

p_value_adjusted_df|>
  filter(p_adjusted==min(p_adjusted))|>
  filter(abs(FC)==max(abs(FC)))|>
  pull(Gene)
```

#### Put any text you may want to type here.

#### End of text block.

#### 2K) Interpretation.

1.  Explain what it means to control the type I error of your p-values
    using a false discovery rate. [1]\
2.  Find the name of your most important gene and describe its function.
    [1]\

#### Put any text for interpretation 2k here. (fixed) \* CHANGED\*

1)  Controlling type I error using the false discovery rate (FDR) means
    limiting the expected proportion of false positives among all genes
    declared significant ,rather than controlling the probability of
    making at least one false positive error.

In genome-wide expression analyses, where thousands of statistical tests
are performed simultaneously,some false positives are unavoidable. FDR
control offers a more appropriate correction than other methods such as
Bonferroni, allowing the identification of truly differentially
expressed genes while keeping the proportion of false discoveries at an
acceptable level. Hence, using FDR ensures a balance between statistical
reliability and biological relevance in large-scale genomic analyses.

2)  The most important gene (based on lowest p-value first and highest
    absolute fold change second) is called Gpr4 and is hypothesised to
    function as a pH signaling protein. Through activation of G Protein
    coupled receptors (GPCRs), it mediates the cellular response to
    extracellular acidity, and is thought to be involved in the
    development of blood vessels . This would explain its upregulation
    in the training cohort, corroborating the statistical results above
    ([RGD (Gpr4
    gene)](https://rgd.mcw.edu/rgdweb/report/gene/main.html?id=1311604 "Gpr4 gene annotation in the Rat Genome Database")).\

#### This concludes Assignment 2.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session.

### Assignment 3: Principal Component Analysis (Gustavo Stolf Jeuken, 8-01-2026)

During this Assignment, you will perform Principal Component Analysis on
both the gene expression and metabolomics data containing all
conditions. You will investigate groupings in the score plots, and have
a look at important variables distinguishing male and female rats, as
well as features that can distinguish the different training durations.

#### 3A)

0.  Load the Metabolomics_Exp and Metabolomics_Values data again and
    meancenter your data.
1.  Perform a principal component analysis on the meancentered
    metabolomics data using the svd() function. [1]\
2.  Make a scatter plot of the scores using the first and second
    principal component. Provide the scores of each treatment group in a
    different color. [1]\
3.  Calculate the variance explained for each component. [0.5]
4.  Show the variance explained of each component in your plot labels.
    [0.5]\
5.  Add a legend to identify the colour for each condition using the
    legend() function. [0.5]\
6.  Also make another scoreplot, similar to the first one, but now plot
    the different sexes in different colors [0.5].

```{r Assignment 3A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggrepel)
#0 
metabolomics_Exp<-read.csv("df_metabolomics_Exp.csv",row.names=1)
metabolomics_Values<-read.csv("df_metabolomics_Values.csv",row.names=1)

#Creating variable for control vs training
metabolomics_Exp<-metabolomics_Exp|>
  separate(Groups,"Treatment"," ",remove = FALSE)|>
  mutate(Treatment=factor(Treatment))
  
# Mean-centering/scaling the data
metabolomics_Values_norm<-apply(metabolomics_Values,2,function(x){(x-mean(x))})

# Running PCA and defining scores & loadings
pca<-svd(metabolomics_Values_norm)
scores= as_tibble(pca$u %*%diag(pca$d))
loadings= as_tibble(pca$v)




#2
# Defining a factor for the groups for the plot colouring
groups=factor(metabolomics_Exp$Groups)

scores$Groups<-groups

# Plotting the first two scores with ggplot2 as it adds a nice legend for 5
score_plot<-ggplot(scores,aes(x=V1,y=V2,colour=Groups))+geom_point()


#3
# Computing total ssq 
totalssq<-sum(metabolomics_Values_norm*metabolomics_Values_norm)

# Computing the explained variance
ssq <- 100 * pca$d^2 / totalssq

cumssq<-cumsum(ssq)
cumssq
round(ssq[1:2],2)


#4
score_plot_grp<-score_plot+labs(x="PC1 (41.1% variance)",y="PC2 (7.2% variance)",
                                title="Score plot of samples for the first to PCs coloured by group")+
  theme(plot.title = element_text(hjust = 0.5))
#5
#Added already by ggplot
#6
  
scores$Sex<-factor(metabolomics_Exp$Rat.Sex, labels=c('Female', 'Male'))

score_plot_sex<-ggplot(scores,aes(x=V1,y=V2,colour=Sex))+
  geom_point()+
  labs(x="PC1 (41.1% variance)",y="PC2 (7.2% variance)",
       title="Score plot of samples for the first to PCs coloured by sex")+
  theme(plot.title = element_text(hjust = 0.5))

score_plot_grp

score_plot_sex
```

#### Put any text you may want to type here.

#### End of text block.

#### 3B: Interpretation.

1.  What is your interpretation? Which source of variation is dominating
    the metabolomics data [1]? Include the percentage of variation
    explained per component in your answer.
2.  Discuss whether there is an effect of training in the metabolomics
    data, and if so, how it can be observed. [1]

#### Put any text for Interpretation 3B here.

1)  There is a clear distinction between male and female rats along PC1,
    which contributes 41.1% of the variance, where the males have a
    higher PC1 value than average and females lower than average.

Along PC2 which contributes 7.2% of the variance, we observe a decrease
in PC2 as time progresses in the training cohort. both in the male and
female population. We further note that the control subjects also fall
on the lower end of the PC2 spectrum, around the value of the rats that
have trained 2-4 weeks.

All in all, the dominant source of variation appears to be the sex of
the rat.

2)  One might speculate that initiation of the training regimen induces
    a positive change in metabolites along the PC2 axis, which
    stabilizes to a lower level after 8 weeks. Without data on the
    controls at different time points, it is impossible to check whether
    the gradient observed in the training cohort is an effect of age,
    and whether the decrease in PC2 would occur regardless of training.

The substantial overlap between control and training samples combined
with the relatively low explained variance of 7.2% indicates that
training does not induce a strong global shift in the metabolomics
profiles. Any training-related effects are therefore likely limited to a
small subset of metabolites and do not contribute substantially to the
major sources of variation captured by the first two principal
components.

#### End of text block.

#### 3C: Loading plot.

The loadings of the PCA link the variation found in between the rats to
the metabolites being measured. To circumvent a very crowded loading
plot, only plot the metabolite names of the 30 most important loadings.
$Importance_j =\sqrt(p_{1j}^2 + p_{2j}^2)$. 1: Calculate importance of
each metabolite.[1] 2: Select the 30 metabolites with highest
importance.[2] 1. Make a scatter plot with the names of the important
metabolites.[1] 2. Add again the percentage explained per PC along the
axes.

```{r Assignment 3C Loading plot, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

#1 

loadings_imp<-loadings|>
  mutate(Importance=sqrt(loadings[,1]^2+loadings[,2]^2),
         Metabolite=colnames(metabolomics_Values))
#2
loadings_imp<-loadings_imp|>
  arrange(desc(Importance))|>
  slice(1:30)

#3
library(ggrepel)


ggplot(loadings_imp,aes(x=V1,y=V2,label=Metabolite))+geom_jitter()+geom_text_repel(size = 3)+labs(x="PC1 (41.1% variance)",y="PC2 (7.2% variance)",title="PCA loadings plot (30 metabolites)")+theme(plot.title = element_text(hjust = 0.5))
```

#### 3D) Interpretation (fixed) *CHANGED*

1.  From the loading plot select some metabolites that are affected by
    training. Discuss why these metabolites could be affected [1].
2.  From the loading plot select some metabolites that are affected by
    Sex. Discuss why these metabolites could be affected.[1]

#### Name some metabolites and give an explanation

1)  As noted above, we observed that training had the most effect along
    the PC2 variable, with increasing training time corresponding to a
    decrease in PC2. The metabolites showing relatively low loadings on
    PC2 include serotonin, taurochenodeoxycholate, hyodeoxycholate and
    sphingosine. These metabolites therefore seem to increase with
    training time. These compounds are biologically plausible candidates
    for training-related effects. Physical training is known to
    influence energy metabolism, lipid signaling and gut–liver axis
    activity, which can affect bile acid composition and sphingolipid
    metabolism[@starnes2017]. In addition, serotonin plays a role in
    metabolic regulation and exercise-induced adaptations, particularly
    through its involvement in energy balance and peripheral signaling
    [@pietrelli2018] .

The presence of these metabolites among the most influential
contributors to PC2 is consistent with a modest but biologically
meaningful metabolic response to training, even though training does not
dominate the overall variance structure of the data.

2)  Metabolites with high loadings on PC1 include cholate, glycocholate,
    sphingomyelin and several triglyceride species. Since males tend to
    score higher along PC1 in comparison to females, this suggests that
    these metabolites occur more prevalently in males. These metabolites
    are closely linked to lipid metabolism, bile acid synthesis and
    lipid transport, processes that are well known to differ between
    males and females due to hormonal regulation[@quiróscognuck2020].
    Sex-specific differences in steroid hormones influence bile acid
    pools and lipid composition, which can explain the strong separation
    observed along PC1[@dimitriadis2019]. The prominence of
    triglycerides and sphingolipids among these metabolites supports the
    conclusion that sex-related biological differences are the dominant
    drivers of variation in the metabolomics data.

#### End of text block.

#### 3E)

1.  Reconstruct your metabolomics dataset using only the first and
    second PCs. [1]\
2.  Calculate the residual matrix of the metabolomics data by
    subtracting the reconstructed data from the original data. [0.5]\
3.  Calculate the Sum of squares of the residuals by first squaring all
    values in the residual matrix and then summing all these values [1]\
4.  Show that the explained variation and the residual variation add up
    to the total variation of the original data [.5]\

```{r Assignment 3E, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

scores<-scores[1:50] # Removing the group and sex variable used for plotting

reconstructed_metabolomics<-as.matrix(scores[,1:2])%*%t(as.matrix(loadings[,1:2])) #Reconstructing matrix

residuals<-metabolomics_Values_norm-reconstructed_metabolomics #Computing residuals

res_ssq<-sum(residuals^2) #Computing the residual ssq

exp_ssq<- sum(pca$d[1:2]^2)
# Computing the absolute explained ssq (not a percentage as above)

round(res_ssq+exp_ssq-totalssq, 4) # comparing sum of explained and residual ssq to the total ssq, difference is 0 showing they are equal
```

#### 3F)

1.  Calculate the sum of of the squared residuals over all rows and make
    a bar chart of these sums. Indicate the female rats and male rats
    with different color to show whether there is a Sex difference in
    residual variation, and add a legend. [1]\
2.  Calculate the sum of of the squared residuals over all columns and
    make a bar chart of these sums.[1]\

```{r Assignment 3F, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
#1
squared_residual_samples<-data.frame("Sex"=factor(metabolomics_Exp$Rat.Sex,label=c("female","male")), 
                                     "Name"=row.names(metabolomics_Exp),
                                    "ssq"=rowSums(residuals^2))

ggplot(squared_residual_samples, aes(x=Name,y=ssq,fill=Sex))+
  geom_col()+
  coord_flip()+
  theme(axis.text.y=element_text(size = 5),
        plot.title = element_text(hjust = 0.5))+
  labs(x="Sample ID",y="Residual sum of squares",title="Barplot of the residual sum of squares per individual")

#2
squared_residual_metab<-data.frame("Metabolite"= colnames(residuals),"ssq"=colSums(residuals^2))

ggplot(squared_residual_metab, aes(x = Metabolite, y = ssq)) +
  geom_col()+
  labs(
    x = "Metabolite",
    y = "Residual sum of squares",
    title = "Barplot of the residual sum of squares per metabolite"
  ) +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )

```

#### Put any text you may want to type here.

#### End of text block.

#### 3G

1.  Which rat has the highest residual variation? [1]

2.  Which variable has the highest contribution to the residuals for
    this rat? [1]\

3.  If we define that the variable with the highest % explained
    variation is the best modeled variable, then which variable has been
    modeled the best in the PCA model with 2 PCs? [1]\

#### Put any text you may want to type here.

We find that the rat with ID 90254 has the greatest residual variation.
For this rat the edetate.1 metabolite is found to be the biggest
contributor to the unexplained variance.

If we define the best modeled variable as the one with the highest %
explained variation, we find that our model with two PCs has
PC.O.32.1.\_feature3 as the best modeled variable. It has an explained
sum of squares of 96.9%, thus being modeled very well.

```{r 3G, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
squared_residual_samples|>
  arrange(desc(ssq))|>
  select(-Name)|>
  slice(1)
# Find the rat with the highest residual ssq

sort(abs(residuals["90254",]),decreasing = TRUE)[1] # Find the metabolite with the highest contribution to the residual ssq for this rat

## Compute variance per metabolite and residual variance
total_ssq_per_metabolite<-colSums(metabolomics_Values_norm^2)
res_ssq_per_metabolite<-colSums(residuals^2)

percent_ssq_metabolite<-sort(100*(total_ssq_per_metabolite- res_ssq_per_metabolite)/total_ssq_per_metabolite,decreasing = TRUE)

percent_ssq_metabolite[1]

```

#### That concludes week 1 of the project!

Please hand in your .rmd and pdf files on Canvas for your pass/fail
grade and feedback. The week 1 deadline is Tuesday January 13, 8:00 (AM)

### Assignment 4: Clustering (Gustavo Stolf Jeuken, 13-01-2026)

In this assignment, you will focus on clustering your metabolomics data.
This will be done using hierarchical clustering, kmeans and UMAP
methods.\

In the code block below, some functions are defined for you to use in
subsequent questions. You do not need to understand what happens in this
code. Do not edit the code block!

#### 4A)

Import the metabolomics_Values and metabolomics_Exp data again using the
read.csv() function.

```{r Assignment 4A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
metabolomic_values<-read.csv("df_metabolomics_Values.csv",row.names = 1)
metabolomic_exp<-read.csv("df_metabolomics_Exp.csv",row.names = 1)
```

#### 4B)

1.  Meancenter and scale the metabolomics data before clustering using
    the scale function. [1]
2.  Calculate the Euclidean distances between the samples in the
    centered and scaled metabolomics expression data using the dist
    function. [0.5]\
3.  Use the hclust function to calculate the hierarchical clustering
    from the Euclidean distance matrix using average linkage.[0.5]\
4.  Plot the hierarchical clustering dendrogram with labels indicating
    Sex [0.5]
5.  5\. Plot the hierarchical clustering dendrogram with labels
    indicating Group [0.5]

```{r Assignment 4B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# standardising the data
metabolomic_values<-apply(metabolomic_values,2,function(x){(x-mean(x))/sd(x)})

#computing euclidean distance between the samples
l2_distance<-dist(metabolomic_values)

l2_hclust<-hclust(l2_distance,method="average")

metabolomic_exp<-metabolomic_exp|>
  mutate(
    Rat.Sex=factor(Rat.Sex,label=c("F","M")),
    Experiment.Group=factor(Experiment.Group)
    )

plot(l2_hclust,labels=metabolomic_exp$Rat.Sex,cex=0.6,
     xlab="Sex",main="Dendogram labeled by sex (average linkage)",ylab="Distance")

plot(l2_hclust,labels=metabolomic_exp$Groups,cex=0.6,
     xlab="Experiment Group",main="Dendogram labeled by experiment group (average linkage)",ylab="Distance")

```

#### Put any text you may want to type here.

Hierarchical clustering was performed on the centered and scaled
metabolomics data using Euclidean distances.The following section
provides an interpretation of the clustering results based on sex and
experimental group, and compares them with the PCA outcome.

#### End of text block.

#### 4C) Interpretation.

1.  Discuss your interpretation of the cluster result. Focus on how well
    the Rats with respect to Sex and Groups are clustered. [1]\
2.  Discuss the differences between the clustering and the PCA results.
    [1]\

#### Put any text for interpretation 4C here.

1)  In the first plot, when samples are labeled by sex, the hierarchical
    clustering shows a clear separation, with samples of the same sex
    clustering together. This indicates that sex is a major source of
    variation in the metabolomics data.

In contrast, when samples are labeled by experimental group, the
clustering does not exhibit a clear separation between Control and
Training groups, although the subjects with 1-2 weeks of physical
training are clustered together. The samples from the remaining groups
are intermixed, suggesting that sex has a stronger effect (explains more
variation) than experimental groups.

2)  Hierarchical clustering and PCS approach the data from different
    perspectives. Hierarchical focuses on pairwise similarities between
    samples and represent them into a dendogram, making it easier to see
    which samples are most similar to each other. Whereas PCA summarizes
    the main sources of variation in the data by projecting samples onto
    a small number of principal components (PC1 and PC2).

In our dataset, both methods lead to consistent conclusions. PCA shows a
clear separation by sex along the first principal component, while
experimental groups largely overlap. Hierarchical clustering reflects
the same pattern, with samples clustering primarily by sex and
secondarily byu training time. However, PCA provides a clearer visual
separation of the dominant effect (Sex), whereas hierarchical clustering
highlights the lack of distinct group-based clusters,whereas
hierarchical cluster requires deciding a number of clusters.

#### End of text block.

#### 4D)

Now perform a hierarchical clustering of the METABOLITES of the
meancentered and scaled version of the metabolomics data using Euclidean
distance and average linkage and plot the dendrogram using the
plot_hclust function. To keep the plots visible only apply this on the
first 36 metabolites, which are amines. [2]\

```{r Assignment 4D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
met_subset<-metabolomic_values[,1:36]

met_dist<-dist(t(met_subset))
met_clust<-hclust(met_dist,method="average")

plot(met_clust,labels = abbreviate(colnames(met_subset),minlength = 7),cex=0.6,main="Dendogram of 36 amine metabolites (average linkage)",ylab="Distance")
```

#### Put any text you may want to type here.

Dendogram interpretation: The hierarchical clustering of the first 36
amine metabolites reveals several well-defined sub-clusters, indicating
groups of metabolites that show similar changes across samples. Closely
related amino acids, such as branched-chain amino acids (e.g. leucine
and isoleucine), cluster together at low Euclidean distances, reflecting
a high degree of similarity in their behaviour. Several metabolites form
tight sub-clusters, suggesting that they vary in a highly similar way
across the dataset, which may point to shared biochemical pathways or
coordinated regulation. In contrast, metabolites that only merge at
higher distances show less similar patterns across samples, indicating
more independent behaviour. \#### End of text block.

#### End of text block.

#### 4E)

1.  Select 2 different clusters of two metabolites (e.g. Isoleucine and
    Leucine, and Ethanolamine and Hydroxylysine) and plot their
    metabolites levels in a scatter plot to explore their similarity.[2]
2.  Discuss the difference in interpretation when clustering samples
    (experimental conditions) or clustering variables (metabolites). [1]

```{r Assignment 4E}
# Select isoleucine and leucine
ggplot(metabolomic_values,aes(x=Isoleucine,y=Leucine))+geom_point()+geom_smooth(method="lm",se=F)+labs(title = "Scatter plot of Leucine vs Isoleucine plasma levels")

# Select alanine and proline
ggplot(metabolomic_values,aes(x=Alanine,y=Proline))+geom_point()+geom_smooth(method="lm",se=F)+labs(title = "Scatter plot of Proline vs Alanine plasma levels")


```

#### Put any text for interpretation of 4E2 here.

Clustering samples is mainly used to understand how experimental
conditions relate to each other and which factors drive differences
between individuals. In this dataset, clustering samples helped reveal
that sex explains a large part of the variation, while training and
control groups are less clearly separated. This type of clustering is
therefore useful for interpreting biological or experimental effects at
the sample level.

Clustering metabolites, on the other hand, focuses on how variables
behave across all samples. It allows the identification of metabolites
that change in a similar way, which can indicate shared biochemical
functions or coordinated regulation. While sample clustering answers
questions about group separation, metabolite clustering provides insight
into relationships between metabolites themselves and underlying
metabolic structure.

#### End of text block.

#### 4F)

1.  Perform a kmeans clustering with 2 clusters and 5 clusters of the
    samples of the metabolomics. [1]\
2.  Make a table of clusters vs experimental groups for both
    clusterings. [1]
3.  Apply kmeans to cluster metabolites. Only select the first 36
    metabolites which are the Amines. Vary the number of clusters based
    on the results in 4D. Show which Amines are grouped in the
    clusters.[1]\

```{r Assignment 4F, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
set.seed(123)
# 2 cluster k-means
k2 <- kmeans(metabolomic_values,2)




table(k2$cluster,abbreviate(metabolomic_exp$Groups,minlength = 14))

table(k2$cluster,metabolomic_exp$Rat.Sex)

set.seed(123)
# 5 cluster k-means
k5 <- kmeans(metabolomic_values,5)

table(k5$cluster,abbreviate(metabolomic_exp$Groups,minlength = 14))

# Amine clustering
set.seed(123)
k_met<-kmeans(t(met_subset),3)


split(rownames(t(met_subset)),k_met$cluster)
```

#### Put any text you may want to type here.

In the hierarchical clustering of the first 36 amines (4D), the
metabolites do not form just one or two large groups. Instead, the
dendrogram shows several clear branches, with 3 main groupings and
smaller sub-groups within them. Because of this structure, using only 2
clusters would be too simplistic and would combine metabolites that show
clearly different patterns. For this reason, we decided on using k=3 for
the clustering.

#### End of text block.

#### 4G) Interpretation.

1.  Summarize the kmeans result. Which experiments are clustered
    together? [1]\
2.  Compare the kmeans cluster result with the PCA result. Discuss
    differences and similarities. [1]\
3.  Compare the hierarchical cluster result of the samples with the
    kmeans result. Discuss differences and similarities. [1]\

#### Put any text you may want to type here.

1)  The k-means clustering of samples shows that experimental groups are
    not cleanly separated into distinct clusters. With k = 2, samples
    from all experimental conditions are distributed across both
    clusters, indicating that the clustering is not driven by group or
    training duration. We checked if the k-means algorithm was
    classifying based on sex, which is shown to be the case in the
    second table.

    Increasing the number of clusters to k = 5 reveals more structure,
    with some clusters containing higher proportions of specific
    training durations. However, control and training samples are still
    mixed across multiple clusters. Overall, the results suggest that
    while k-means captures variability within the data, experimental
    condition alone does not fully explain the clustering patterns.

2)  The k-means clustering and PCA results are largely consistent with
    each other. PCA shows a very clear separation by sex along the first
    principal component, indicating that sex is a major source of
    variation in the metabolomics data. We observe this clustering based
    on sex in the k-means clustering with k=2, where there is a perfect
    sex classification of the subjects.

    When samples are coloured by experimental group, the PCA shows
    overlap between control and training samples, with an increase in
    training time corresponding to a decrease along PC2. We observe that
    subjects with 1-2 weeks of training score significantly higher along
    PC2 in both males and females. While k-means identifies some
    internal structure in the data, it does not produce clusters that
    correspond clearly to experimental conditions, which shows, again,
    the overlap observed in the PCA plots. Clusters 2 and 3 are
    predominantly occupied by the samples undergoing training for 1-2
    weeks. We can also highlight these subjects have been evenly split
    between clusters 2 and 3, suggesting a sex difference.

    Overall, PCA highlights the dominant sources of variation in a
    low-dimensional representation, whereas k-means assigns samples to
    distinct clusters. Both approaches lead to the same conclusion:
    experimental group effects are weaker than sex-related differences
    and do not drive a clear separation of samples. In both methods we
    can observe a separation between 1-2 weeks of training and the other
    conditions.

3)  Hierarchical clustering and k-means show consistent overall patterns
    but differ in how they represent structure in the data. Hierarchical
    clustering highlights progressive similarity relationships between
    samples and reveals that samples tend to cluster more strongly by
    sex than by experimental group . In contrast, k-means forces samples
    into a predefined number of clusters, which results in mixed
    clusters containing samples from multiple experimental conditions.

    While hierarchical clustering provides a clear overview of how
    samples relate to each other at different similarity levels, k-means
    produces a fixed clustering that depends on the chosen number of
    clusters. Despite these methodological differences, both approaches
    lead to the same conclusion: controls and training groups are not
    clearly separated at 8 weeks, although the 1-2 week training cohort
    are clustered together. Moreover, sex-related differences appear to
    be a more dominant source of variation.

#### End of text block.

#### 4H) UMAP.

1.  Apply a UMAP on the samples of the centered and scaled metabolomics
    data. Vary the number of neighbours until you have a "good" result.
    [0.5]\
2.  Apply a UMAP on the metabolites. For simple visualization reasons
    only use the first 36 metabolites, which are amines. Vary the number
    of neighbours until you have a "good" result.[0.5]

Note! Fix the random state such that the same can be repeated!

```{r Assignment 4H, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# Sanple UMAP
set.seed(123)
custom.config.sample <- umap.defaults
custom.config.sample $n_neighbors <- 3

umap.sample<-umap(metabolomic_values,custom.config.sample)

umap.sample.coord<-data.frame(umap.sample$layout)|>
  mutate(group=metabolomic_exp$Groups,
         sex=metabolomic_exp$Rat.Sex)

sampleplot_group<-ggplot(umap.sample.coord,aes(x=X1,y=X2,colour=group))+geom_point()+labs(title="UMAP plot of samples coloured by group (3 neighbours)")

sampleplot_sex<-ggplot(umap.sample.coord,aes(x=X1,y=X2,colour=sex))+geom_point()+labs(title="UMAP plot of samples coloured by sex (3 neighbours)")

sampleplot_group
sampleplot_sex


# Sanple UMAP per sex
set.seed(123)
custom.config.sample <- umap.defaults
custom.config.sample $n_neighbors <- 3 # 3,4 works well


#Extra option for plotting males and females seperately
males<-tibble(metabolomic_values[metabolomic_exp$Rat.Sex=="M",])


females<-tibble(metabolomic_values[metabolomic_exp$Rat.Sex=="F",])

umap.m<-umap(males,custom.config.sample)
umap.m.coord<-data.frame(umap.m$layout)|>
  mutate(Group=metabolomic_exp$Groups[metabolomic_exp$Rat.Sex=="M"])

umap.f<-umap(females,custom.config.sample)
umap.f.coord<-data.frame(umap.f$layout)|>
  mutate(Group=metabolomic_exp$Groups[metabolomic_exp$Rat.Sex=="F"])


mplot<-ggplot(umap.m.coord,aes(x=X1,y=X2,colour=Group))+geom_point()+labs(title = "UMAP plot of male samples")
fplot<-ggplot(umap.f.coord,aes(x=X1,y=X2,colour=Group))+geom_point()+labs(title = "UMAP plot of female samples")

mplot/fplot

# Metabolite UMAP
set.seed(12)
custom.config.met <- umap.defaults
custom.config.met $n_neighbors <- 6

umap.met<-umap(t(met_subset),custom.config.met)

umap.met.coord<-data.frame(umap.met$layout)

metplot<-ggplot(umap.met.coord,aes(x=X1,y=X2))+geom_point()+labs(title="UMAP plot of amine metabolites (6 neighbours)")

#install(patchwork)

metplot



```

#### Put any text you may want to type here.

1)  UMAP is designed to preserve local neighbourhood relationships while
    still providing a meaningful global representation of the data. With
    very small numbers of neighbours, the embedding focuses on local
    relationships, which leads to very tight and exagerated clusters.
    With larger values, local structure becomes less dominant and the
    separation between groups starts to clearly separated. For the
    sample clustering, 3 neighbours ended up matching expectations from
    PCA and the previous clustering algorithms the best.

2)  For the metabolites, using 6 neighbours provides a good balance
    between both extremes: metabolites that behave similarly are placed
    close together, while the overall structure remains easy to
    interpret. For this reason, the UMAP result with 6 neighbours was
    selected. There is also correspondence between metabolites that lie
    closely together in the UMAP, and those that are clustered together
    in the hierarchical clustering.

#### End of text block.

#### 4I) Interpretation.

1.) What is your interpretation of the UMAP clusters of the metabolite
samples.[1]\
2.) What is your interpretation of the UMAP clusters of the metabolites
in comparison with kmeans and hierarchical clustering.[1]\
3.) Discuss the differences between UMAP and k-means or hierarchical
clustering. [1]\

#### Put any text you may want to type here.

1.   The UMAP representation of the samples shows a very clear
    separation by sex, with male and female samples forming two
    well-defined groups. This indicates that sex is a major source of
    variation in the metabolomics data. Furthermore, within both sexes
    we observe clustering, although there is no clear distinction
    between control and treatment group. We do observe a clear cluster
    subjects with 1-2 weeks training, particularly in the female
    population. The fact that the clusters do contain a mixture of
    treatment groups suggests that training status or duration has a
    weaker effect on the overall metabolic profile. Overall, the UMAP
    results confirm that sex-related differences dominate the sample
    structure.

<!-- -->

2.  The UMAP analysis of the amine metabolites reveals groupings that
    are consistent with both k-means and hierarchical clustering.
    Metabolites that were grouped together in the dendrogram and in
    k-means clustering also tend to appear close to each other in the
    UMAP plot. This suggests that these metabolites share similar
    patterns across samples. While UMAP does not assign metabolites to
    strict clusters, it provides a complementary view that supports the
    same overall relationships identified by the other clustering
    methods. This suggest a entangled relationship between the
    metabolites rather than a discrete clustering.

<!-- -->

3.  UMAP differs from k-means and hierarchical clustering in that it is
    primarily a visualisation method rather than a clustering algorithm.
    It focuses on representing similarities between points in a
    low-dimensional space, making patterns and groupings easier to
    observe. In contrast, k-means and hierarchical clustering explicitly
    assign samples or metabolites to clusters. In this dataset, UMAP is
    particularly useful for highlighting dominant patterns, such as the
    strong separation by sex.

#### End of text block.

#### This concludes Assignment 4.

Make sure you save your .Rmd file and workspace so you can continue in
your next project session.

### Assignment 5: Classification (Johan Westerhuis, 15-01-2023)

In this assignment we will use the LDA and PCDA methods to find the
metabolites that best discriminate between Control and 8 weeks training
conditions.\

#### 5A)

0)  Load the df_metabolomics_Values and the df_metabolomics_Exp
    datasets.
1)  Make a reduced metabolomics dataset by selecting only the samples of
    Training 8 weeks and Control 8 weeks. [1]\
2)  For visualization purposes, only select metabolites 80 (CAR(10:0-OH)
    until metabolite 613 (TG 60:4). These are lipids obtained with
    reversed phase liquid chromatography positive mode.[1]
3)  Center and scale this data set again. [1]\
4)  Create a vector of binary values encoding the condition of each
    sample (0 = Control 8 weeks and 1 = Training 8 weeks). [1]\

```{r Assignment 5A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# 0
metabolomic_values<-read.csv("df_metabolomics_Values.csv",row.names = 1)
metabolomic_exp<-read.csv("df_metabolomics_Exp.csv",row.names = 1)

# 1
reduced_idx <- which(metabolomic_exp$Experiment.Time == "8 weeks")

reduced_values<-metabolomic_values[reduced_idx,]
reduced_exp<-metabolomic_exp[reduced_idx,]
   
# 2 
reduced_values<-reduced_values|>
  dplyr::select(80:613)
# 3      
reduced_values<-as.data.frame(scale(reduced_values))


# 4
condition<-reduced_exp|>
  mutate(condition=factor(ifelse(Experiment.Group=="Control",0,1),labels=c("Control","Treatment")))|>
  pull(condition)


```

#### Put any text you may want to type here.

#### End of text block.

#### 5B)

Perform a linear discriminant analysis on your new metabolomics dataset,
using your new vector of binary values as the class of each sample. [2]

```{r Assignment 5B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
lda_results<-lda(reduced_values,condition)

d_lda<-lda_results$scaling


```

#### Put any text you may want to type here.

#### End of text block.

#### 5C) Interpretation.

You should have gotten a warning from your LDA function. Explain this
warning. [2]\

#### Put any text for Interpretation 5c here.

The warning says that "variables are collinear." This indicates that
some variables are highly correlated and can be written as linear
combinations of other variables. This is expected in lipid metabolomics
data for two reasons. In lipid metabolomics many metabolites belong to
the same biochemical pathways or lipid classes and therefore show
similar behavior across samples. Moreover, this type of data usually has
many more variables than observations. These factors causes the
covariance matrix to not be full rank, leading to collinearity.

Collinearity does not prevent LDA from running, but it means that the
discriminant direction is not unique and may depend on a subset of
correlated variables.

#### End of text block.

#### 5D)

To solve the problem being discussed in 5C, you will make a PCDA model
instead of an LDA model.

1\. Apply Principal Component Analysis using the svd function and select
three components. [2]\
2. Make scoreplots of PC1 vs PC2 and PC2 vs PC3 in which the samples are
colored based on their group. [1]\
3. Which of the PCs seems most useful for the discrimination between the
two groups? Think of a way to quantify this. [2]

```{r Assignment 5D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# PCA
met_svd<-svd(reduced_values,nu=3,nv=3)
scores = as.data.frame(met_svd$u %*%diag(met_svd$d[1:3]))
loadings = as.data.frame(met_svd$v)

scores_plot<-scores|>
  mutate(Group=condition)

totalssq<-sum(reduced_values^2)
ssq_explained <- 100 * met_svd$d^2 / totalssq
# scoreplot PC1 vs PC2
PC12<-ggplot(scores_plot,aes(x=V1,y=V2,colour=Group))+geom_point()+labs(x="PC1 (50.2% variance)",y="PC2 (12.6% variance)",title="PC1 vs PC2")+theme(plot.title = element_text(hjust = 0.5))
# scoreplot PC2 vs PC3
PC23<-ggplot(scores_plot,aes(x=V2,y=V3,colour=Group))+geom_point()+labs(x="PC2 (12.6% variance)",y="PC3 (7.8% variance)",title="PC2 vs PC3")+theme(plot.title = element_text(hjust = 0.5))

# Plot together 
PC12+PC23+plot_layout(guides = "collect") &
  theme(legend.position = "right")


#3

mean_ctr<-apply(scores[condition=="Control",],2,mean)
mean_trt<-apply(scores[condition=="Treatment",],2, mean)

mean_diff<-mean_ctr-mean_trt

p_values=apply(scores[condition=="Control",]-scores[condition=="Treatment",],2, function(x) {"p-value"=t.test(x)$p.value})

data.frame(mean=mean_diff,"p-value"=p_values)

# We visually see the biggest difference between control and treatment is along 
# PC3. We confirm this by computing the difference of means between the treatment groups per principle component.The results corroborate the visual observation that PC3 is the most useful component for classification. 
```

#### 5D Interpretation:

We visually see that the PC2 vs PC3 plot reveals a clear separation
along the PC3 axis. From the two groups tend to occupy different regions
of the plot, suggesting that PC3 captures variation related to the
experimental conditions.

We confirm this by computing the difference of means between the
treatment groups per principle component. We observe that PC3 has the
biggest mean difference between the conditions (7.55), as well as the
lowest p-value (0.02). These results corroborate the visual observation
that PC3 is the most useful component for classification.

#### 5E)

1.  Make a PCDA model using the PCA scores and the vector of binary
    class labels .[1]\
2.  Calculate the PCDA regression coefficients of each variable.[1]\
3.  What are the names of the 5 most important metabolites based on
    their absolute PCDA regression coefficients. [1]\
4.  Use the PCDA regression coefficients to make a prediction for each
    rat whether they belong to the control or training group.[1] 5. How
    many misclassifications did you make? [1]

```{r Assignment 5E, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# Run lda on first 3 PCs (PCDA)
PCDA_res<-lda(scores,condition)

# Compute the discriminant vector and the regression coefficients for each metabolite
d_PCDA<-PCDA_res$scaling
coef_PCDA<-as.matrix(loadings)%*%d_PCDA

names(coef_PCDA)<-colnames(reduced_values)


# Find most important metabolites
names(sort(abs(coef_PCDA),decreasing = T)[1:5])

#Compute prediction
pred<-factor(ifelse(as.matrix(reduced_values)%*%coef_PCDA<0,"Control",'Treatment'))

# Compute no. misclassifications
data.frame(condition,pred)[condition!=pred,] # 3 misclassification

table(Truth=condition,Prediction=pred)

```

#### How many misclassifications did you make?

We find that the most important 5 metabolites in descending order of
effect size are LPC.18.2.0.0.\_feature2, PC.30.0.\_feature2,
LPC.20.3.\_feature1and2, LPC.17.1.0.0.\_feature2 and PC.35.3.\_feature1.

The PCDA classifier model misclassifies 3 samples: we see in the
confusion 1 control gets classified as being in the treatment group and
2 treatment group samples get mistakenly classified as a control sample,
as can be seen in the confusion matrix. Counting the treatment group as
positive and the controls as negative, we then get a false positive rate
of 10% and a false negative rate of 20%.

#### End of text block.

#### 5F) Interpretation.

1.  Discuss the similarities and differences between the PCA and PCDA.
    [1]\
2.  Discuss the difference between using t-test to make a Volcano plot
    and using PCDA to calculate important metabolites [1].

#### Put any text for interpretation 5F here.

An important thing to note is that PCA is a step in the process of PCDA.
They are therefore intimately related, although the PCA on its own
serves a very different purpose: it finds direction vectors that explain
the maximal amount of variance in the sample space. PCDA, on the other
hand, uses PCA as a means of dimensionality reduction, to then perform
an LDA in the lower dimensional space of the first $n$ PCs. This then
yields a discriminant vector, that is a vector $d$ maximising the ratio
$\frac{d^{\intercal}Bd}{d^{\intercal}Wd}$, where $B$ and $W$ are the
between group and within group covariance matrices, respectively. This
can be interpreted as maximising the level of separation of the two
groups.

When performing a multitude of t-tests to make a Volcano plot, we are
testing the effects of each metabolite (or variable in general) on the
response separately. Contrarily, in PCDA and LDA the discriminant vector
is a linear combination of the metabolite concentrations detected, and
the weights of this combination indicate the effect of that metabolite.
This is an important difference for a number of reasons. Firstly, the
Volcano plot approach suffers from the multiple testing problem and
requires corrections (such as FDR control) accordingly, while PCDA only
requires one test. Secondly, it is possible that no single metabolite
has a very strong effect, while a combination of metabolite levels has
much stronger predictive power. In this case there will be few important
metabolites in the Volcano plot, while the discriminant vector can pick
up on this combined effect.

#### End of text block.

### Assignment 6: ASCA (Johan Westerhuis, 16-01-2026)

For ASCA we will use the HDANOVA and the mixlm packages. Obtain the
latest version from CRAN.

```{r}
#install.packages("mixlm")
#install.packages("HDANOVA") 

library(mixlm)
library(HDANOVA)
```

Load the metabolomics_Values and the metabolomics_Exp files again on
which the ASCA model will be applied. Use the same set of features as in
the Classification assignment.

#### 6A)

Apply preprocessing of your data. Consider transformation, centering
and/or scaling the data. Make plots of the data (before and after the
preprocessing steps) to defend your choices. [3]

#### Add code for figures here

```{r Assignment 6A, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}

library(tidyr)
library(dplyr)
library(ggplot2)
library(patchwork)
library(ggrepel)

## Loading the data
metabolomic_values<-read.csv("df_metabolomics_Values.csv",row.names = 1)
metabolomic_exp<-read.csv("df_metabolomics_Exp.csv",row.names = 1)

# Selecting lipids
reduced_values<-metabolomic_values|>
  dplyr::select(80:613)

raw_sd_mean<-data.frame(mean=apply(reduced_values,2,mean),
                        sd=apply(reduced_values,2,sd))

set.seed=12
sel<-sample(seq_len(ncol(reduced_values)),25)

sel_data<-reduced_values[,sel]


boxplot(sel_data,las=2,cex=0.6,names = abbreviate(colnames(sel_data)), main="boxplot of 25 random metabolites before processing")

ggplot(raw_sd_mean,aes(x=mean,y=sd))+geom_point()+labs(x="Mean",y="Standard deviation",title="Mean-variance relationship for the lipid metabolome")

# Attempt for square root transform, which did not yield desired results

# sqrt_met<-sqrt(reduced_values-min(reduced_values))
# 
# boxplot(sqrt_met[,sel])
# sqrt_sd_mean<-data.frame(mean=apply(sqrt_met,2,mean),
#                         sd=apply(sqrt_met,2,sd))
# ggplot(sqrt_sd_mean,aes(x=log(mean),y=log(sd)))+geom_point()
# 

processed_metabolomic_values<-as.data.frame(scale(reduced_values))


boxplot(processed_metabolomic_values[,sel],las=2, cex=0,6, names = abbreviate(colnames(sel_data),minlength = 5),main="boxplot of 25 random metabolites after standardisation")




```

#### Add arguments here

We begin by visualising the data with a boxplot of 25 randomly selected
metabolites, along with a mean-standard deviation plot for the entire
data set. The boxplots suggests that the data is distributed fairly
symmetrically, and we observe no clear relation between the mean and
standard deviation, suggesting that the data is homoscedastic. However,
the features do differ greatly in their spread and center. For this
reason we opted to limit the data preprocessing to centering and scaling
the data. We then observe that the data are centered at 0 and exhibit
similar interquartile distances in the postprocessing boxplot.

#### 6B) Data handling

Make the data from wide to long format that is necessary for the asca
function in the HDANOVA package. Make sure the factors Sex and Group are
defined as factors. (See ASCA exercise) [2]

```{r Assignment 6B, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
long_met_data=data.frame(sex=factor(metabolomic_exp$Rat.Sex,labels=c("F","M")), 
                         group=factor(metabolomic_exp$Groups),
                         metab=I(as.matrix(processed_metabolomic_values)))


```

#### 6C) ASCA model:

1)  Create an ASCA model using the only the main factors Sex and Group.
    Do not consider interaction terms yet. [1]
2)  Explain the meaning of Sum.Sq and Expl.var. from the ASCA table, and
    explain how they are calculated. [1]
3)  What is your interpretation of the ASCA model? [1]

```{r Assignment 6C, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
asca_mod<-asca(metab~sex+group,data=long_met_data)

summary(asca_mod)
```

#### Give explanation and interpretation here

2)  In the ASCA table ,"Sum.Sq." represents the sum of squares
    associated with each factor in the model and quantifies how much of
    the total variability in the metabolomics data is explained by that
    factor. It is calculated by decomposing the total sum of squares of
    the data into contributions from each experimental factor and the
    residuals.

    Explained variance(Expl.var.) is the percentage of the total sum of
    squares that is explained by each factor. It is obtained by dividing
    the sum of squares of a given factor by the total sum of squares and
    multiplying by 100. This value indicates the relative importance of
    each factor in explaining the overall variability in the dataset.

3)  The ASCA results show that Sex explains a substantial proportion of
    the variability in the metabolomics data (45.22%), indicating strong
    systematic differences in metabolite profiles between sexes. In
    contrast, the Group factor (Control vs Training at 8 weeks) explains
    only a small fraction of the total variance (12.80%), suggesting
    that the training intervention has a relatively minor effect on the
    metabolomic profiles at this time point. A large proportion of the
    variance remains in the residuals (41.98%), which reflects
    individual variability and sources of variation not captured by the
    experimental factors included in the model. Overall, the ASCA model
    indicates that sex-related differences dominate the structured
    variation in the data, while the effect of training is comparatively
    weak.

#### End of interpretation

#### 6D)

1.  Show score and loading plots of PC1 vs PC2 of the Group factor.
    [0.5]
2.  Show the loading plots of PC1 vs PC2 of the Group factor. Only show
    the 20 most important variables where
    $Importance_j =\sqrt(p_{1j}^2 + p_{2j}^2)$. [1.5]
3.  Give your interpretation of the score and loading plots. [1]

```{r Assignment 6D, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}


scoreplot(asca_mod,factor="group",main="Score plot of the group factor",legend="bottomleft",projections = T)


loadingplot(asca_mod,factor="group",scatter=T,main="Loading plot of the group factor")              

load<-loadings(asca_mod,factor = "group")

Importance=sqrt(load[,1]^2+load[,2]^2)

idx<-which(Importance%in% sort(Importance,decreasing = T)[1:20])

load<-as.data.frame(load[idx,])
load$names<-rownames(load)


ggplot(load,aes(`Comp 1`, `Comp 2`,label=abbreviate(names)))+geom_point(size=0.5)+geom_text_repel(size=3,max.overlaps = 12)+labs(x='PC1 (64%)', y='PC2 (20%)',title="Loading plot of the 20 most important metabolites")


```

#### Give interpretation of 6D3 here

The score plot for the Group factor shows that most of the group-related
variation is captured by PC1 (64%), indicating that the main differences
between groups are driven by a single dominant component. Along this
axis, there is a clear positive shift from the control group towards the
1-2 weeks training groups. Interestingly, later training time points are
barely shifted relative to the controls, with the 8 weeks training group
even showing a negative shift. This suggests a metabolomic effect that
is time related, and on which the impact of training is unclear.

PC2 (20%) explains a smaller proportion of the variance and mainly
reflects within-group variability, as there is considerable overlap
between groups along this axis. We do observe that at 8 weeks time, the
control group and training group are fairly separated along this axis,
while intermediate time points in the training cohort overlap along the
PC2 axis and are therefore not separated.

The loading plot shows that the group-related separation is driven by a
limited number of metabolites, while most variables contribute only
weakly. The plot of the 20 most important metabolites confirms that
these compounds have the largest influence on the separation observed in
the score plot, particularly along PC1. Metabolites with similar loading
directions are positively associated, whereas metabolites loading in
opposite directions contribute to differences between early and late
training stages/conditions.

Overall, the score and loading plots indicate that the group effect is
present but moderate, with training inducing consistent but relatively
subtle changes in the metabolomic profile. We again observe that sex is
the dominant source of variance. This interpretation is in agreement
with the ASCA results, where the group factor explains a smaller
fraction of the total variance compared to other sources of variation.

#### 6E) Add the two-factor interaction Sex:Group to your model.

1.  Make a new model in which the two-factor interaction Sex:Group is
    also included in the model.[1]
2.  What is your conclusion of this ASCA model with interaction? [1]

```{r Assignment 6E ASCA permutation analysis, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
int_mod<-asca(metab~sex*group,data=long_met_data)

summary(int_mod)
```

#### Give your conclusion of the interaction model

Including the Sex:Group interaction in the ASCA model shows that sex
remains the dominant factor influencing the metabolomic profiles,
explaining the largest proportion of the total variance. The training
effect (Group) alone explains a much smaller fraction of the
variability, indicating that the overall impact of training at 8 weeks
is modest. The interaction term Sex:Group explains 7.72% of variance,
suggesting that the metabolic response to training differs slightly
between males and females. Although this interaction effect is smaller
than the main effects, it suggests the presence of sex-dependent
training responses rather than a uniform effect across sexes. A
substantial proportion of the variance remains in the residuals
(34.26%), reflecting biological variability and other sources of
variation not captured by the experimental factors included in the
model. Overall, the ASCA model with interaction indicates that while sex
is the dominant driver of metabolic variation, training induces modest
effects that partially depend on sex.

#### End of conclusion

#### 6F: Statistical Inference

1.  To conclude whether the main factors and interaction are
    significant, apply a permutation test and plot the permutation
    result using the permutation plot function for both main factors and
    the interaction.[1]
2.  Explain why in ASCA a permutation test is used and not an F-test.[1]
3.  What is your interpretation of the permutation test? [1]

```{r Assignment 6F ASCA permutation analysis, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
mod.perm<-permutation(int_mod)
summary(mod.perm)
par(mfrow=c(1,3))
permutationplot(mod.perm,factor = "sex")
permutationplot(mod.perm,factor = "group")
permutationplot(mod.perm,factor = "sex:group")
```

#### Give explanation of 6F2 here

In ASCA, a permutation test is used instead of a F-test because the
response variable is multivariate and high-dimensional, with many
correlated variables. Metabolomics data typically violate the
assumptions required for an F-test, such as multivariate normality,
independence between variables, and error structure. In addition, the
number of variables often exceeds the number of samples, making
classical parametric tests unreliable.

Permutation testing does not rely on these assumptions. Instead, it
assesses significance by repeatedly permuting the sample labels and
recalculating the explained variance under the null hypothesis of no
association between the experimental factors and the data. This
generates an empirical null distribution, against which the observed
explained variance can be compared. As a result, permutation tests
provide a robust and appropriate way to assess statistical significance
in ASCA.

#### Give interpretation of 6F3 permutation test here

The permutation test shows that all three factors are statistically
significant, with a p-value of 0.00. This indicates that the observed
explained variance for Sex, Group and their interaction term Sex:Group
are much larger than expected under the null hypothesis. This confirms
that these factors have has a significant effect on the metabolomic
profiles.

Overall, the permutation test supports the conclusion that sex is the
main driver of systematic variation in the metabolomics data, with
significant, albeit smaller, effects of training and its interaction
with sex.

#### 

#### 6G)

When exploring the model with interaction, it is common to combine the
interaction effect with one of the main factor effects to improve
interpretation.

1\. Make an ASCA model in which the Sex:Group interaction effect is
combined with the main Group effect.[1]

2.  Make two score plot (using the timeplot function) of the combination
    effect. In the first the PC1 scores of the Females and Males are
    plotted as a function of the Groups. In the second the PC1 scores of
    the Groups are plotted as a function of the Females and Males.[1]

3\. Make a loading plot for PC1 of the combination effect [1].

4\. What is your interpretation of the combination effect (Group +
Sex:Group) [1]

```{r Assignment 6G ASCA Combination model, include=TRUE, eval=TRUE, echo=TRUE, message=FALSE}
# 1

comb_mod<-asca(metab~sex+comb(group+sex:group),long_met_data)
summary(comb_mod)

levs <- levels(long_met_data$sex)
cols <- seq_along(levs)


timeplot(comb_mod,factor = "sex", x_time = T, 
         time="group",
         comb = 2,
         comp = 1,
         xaxt='n',
         main="PC1 score plot as a function of training group per sex",
         ylab="PC1 score",
         xlab = "Sex")

legend("bottomright",
       legend = levs,
       col = cols,
       lwd=2,
       cex = 0.6)

levs <- levels(long_met_data$group)
cols <- seq_along(levs)


timeplot(comb_mod,time='sex',factor="group",comb=2,comp=1,x_time = T,
          main="PC1 score plot as a function of sex per treatment group",
         ylab="PC1 score",
         xlab="Training group")


legend("bottomright",
       legend = levs,
       col = cols,
       lwd=2,
       cex = 0.6)


##
loadingplot(comb_mod,pc=1,factor = "sex",
            main="PC1 loading plot of lipid metabolites",
            xlab="Metabolite",
            ylab="PC1 loading")
        
```

#### Give your interpretation of 6G4 here

The combined effect of Group and Sex:Group explains a substantial
proportion of the structured variation in the metabolomic data (20.51%),
indicating that the training effect cannot be fully interpreted without
considering sex differences.

The timeplots of the PC1 scores show that females and males follow
different response patterns across the training groups, confirming the
presence of a sex-dependent training effect. We observe male subjects
scoring higher along the PC1 axis in the control group and at 8 weeks of
training, while female subjects score higher in the other groups. This
could also suggest an aging related phenomenon, given the similarity
between subjects after 8 weeks.

Across groups, the general trend along PC1 suggests that training
induces systematic changes in the metabolomic profiles, but the
magnitude and direction of these changes differ between sexes. While
both females and males show consistent trends across training duration,
the separation between groups is more pronounced for some time points,
indicating that sex modulates the metabolic response to training rather
than simply shifting all profiles in the same way.

The second timeplot, where PC1 scores are shown as a function of sex for
each group, further supports this interpretation by showing that group
differences are not constant across sexes. This confirms that the
interaction between sex and group contributes meaningfully to the
observed variation.

The loading plot of PC1 shows that the combined effect is driven by a
subset of metabolites with higher loadings, while most metabolites
contribute only weakly. This indicates that the sex-dependent training
response is associated with coordinated changes in a limited number of
metabolites rather than a global shift of the metabolome.

Overall, the combined ASCA model reveals that training induces moderate
but structured metabolic changes that depend on sex, and that combining
the main Group effect with the Sex:Group interaction provides a clearer
and more biologically meaningful interpretation of the training
response.

#### This concludes assignment 6

Make sure you save your .Rmd file and workspace so you can continue in
your next project session.

#### That concludes week 2 of the project!

Please hand in your .rmd and .pdf files on Canvas for your project
grade. The week 2 deadline is Tuesday January 20, 08:00 (AM).\*\*\*
