---
title: "R TUTORIAL"
subtitle: Working with Vectors and Matrices in R
output:
  pdf_document: default
  html_document:
    df_print: paged
    toc: no
    toc_depth: 4
---

## Preface

The main goal of this first exercise is to work with vectors and matrices. Learn how they can be created, how vectors and matrices can be added or multiplied to each other, how to divide a matrix by another matrix etcetera. To make linear algebra useful in high dimensional data analysis it is important to realize that these vectors and matrices represent measurements. These measurements come in datasets such as glucose levels in plasma of 100 diabetics and 100 controls, the expression of 20,000 genes from Arabidopsis Thaliana obtained from TAIR, or 8000 OTUs (Operational Taxonomic Units) measured of poo samples of 800 monkeys in different areas in Africa. If you multiply such a dataset with another vector, the result has a specific meaning in terms of the samples, individuals or the features being measured.

Most of you will have followed the Statistics with R course and will be familiar with most topics in this R tutorial. In that case you can skip most parts of this tutorial, except the part on Matrices in R. Make sure you understand the Linear Algebra in R part.

## Vectors

Vectors are the simplest data type in R.

### Defining a vector

The following creates a column vector which has the name `v`. The first entry in the vector is a 3 and the second entry is a 1. (*Don't forget to run the chunk*.)

```{r}
v <- c(3,1) 
```

In the global environment (one of the tabs in the upper right pane in RStudio), you can see the data type of the elements (`num` for *numeric*), the dimension of the vector, and its values. If you want R to output the values the you can just type:

```{r}
v
```

You can define a vector of any size in this manner:

```{r}
v <- c(3, 1, 7, -21, 5, 6)
v
```

Notice, though, that this always creates a column vector even though it is printed "in row format". If you wanted to create a row vector (sometimes indicated in equations by a superscript: $\mathbf{v}^\mathrm{T}$ ), you need to take the transpose of a column vector. The transpose is defined using the `t()` function:

```{r}
vt <- t(v)
vt
```

A common task is to create a large vector with numbers that fit a repetitive pattern. R can define a set of numbers with a common increment of 1 using colons. For example, to define a vector those first entry is 1, the second entry is 2, the third is 3, up to 8 you enter the following:

```{r}
v <- 1:8
v
```

*How can you get the numbers to count down?*

If you wish to use an increment other than 1, then you can use a function ( `seq()` ), where the start number, the last number and the value of the increment, are needed. For example, to define a vector that starts with 2 and ends in 4 with steps of .25 you enter the following:

```{r}
v3 <- seq(from = 2, to = 4, by = 0.25)
v3
```

### Some more notes on R

#### Built-in functions

*You have seen two functions in R now:* `t()` *and* `seq()`.

*Most functions need an argument as input, e.g.:*

```{r}
X <- mean(v3) # calculates the mean of all elements in v3
Y <- rnorm(10) # produces 10 normally distributed random numbers
```

*To obtain information on the use of any function, you can type* `?` *and the function name, e.g.,* `?seq` *into the console or a code block in the Notebook, or you can use the search function in the* "Help" *tab in the bottom right pane in RStudio.*

```{r}
?seq
?c
```

*We'll look more closely at custom-made functions later.*

#### Comments

The `#` is used to add comments in the code (see two chunks above).

#### Workspace

The variables that have been defined above (`v`, `vt`, `v3`) are stored in R's *workspace*. You can see them in the Environment window. If you want to clear all variables from the workspace, click the small broom. Remove single objects with the `rm()` function:

```{r}
rm(v3)
```

### Accessing elements within a vector

You can view individual entries in a vector asking for a subset. For example, to view the first entry, just type in the following:

```{r}
v[1]
```

*Note: in R, indices are counted from 1 (not 0 which is used in most programming languages).*

This gives the first 3 elements of the vector.

```{r}
v[1:3]
```

*Test what happens, if you ask for elements of a vector that are not defined.*

To get all elements of a vector, except one, use for example:

```{r}
v[-3]
```

You can also access elements of a vector using a Boolean vector:

```{r}
smallerThan5 <- 1:length(v) < 5
smallerThan5
v[smallerThan5]
```

*Note, this has troubles with `NA` values, so it's safer to convert the Boolean vector into vector of indices using* `which()`:

```{r}
which(smallerThan5)
v[which(smallerThan5)]
```

#### Vectors of repeated elements

You can create vectors filled with all the same number using the `rep()` function:

```{r}
z0 <- rep(0, times=4)
z1 <- rep(1, times=6)

z0
z1
```

### Basic operations on vectors

#### Addition and subtraction

R will add vectors by pairwise addition of the elements in each:

```{r}
v <- 1:5
u <- 2:6
v+u
```

```{r}
u1 <- 1:10 
```

`u1` *is a name so it is different from* `u[1]` *and from* `u * 1.`

```{r}
v + u1
```

In mathematics (and also many other programming languages) this last addition is not possible because the size of $\mathbf{v}$ and $\mathbf{u1}$ are different. *Try to find out how R handles this.*

#### Multiplication with a scalar

You can multiply each element with the same number, like so:

```{r}
-2 * v
```

*R actually doesn't usually use scalars, but thinks of them as vectors with just one element. Don't worry about this now.*

```{r}
v/3
v3 = (8:16)/4
v3
```

*Remember the vector* `v3` *from above?* H*ow could you define the same vector using the colon notation?*

#### Inner product

The *inner product also called the dot product* of two vectors $\mathbf{a} = (a_1, a_2, ..., a_n)$ and $\mathbf{b} = (b_1, b_2, ..., b_n)$ is defined as $\mathbf{a}·\mathbf{b} = a_1b­_1 + a_2b_2 + ... + a_nb_n$ .

```{r}
a <- c(1,3,2)
b <- c(2,8,9)
sum(a*b)
```

This is a scalar.

If the two vectors are *perpendicular* or *orthogonal*, the inner product is equal to 0.

```{r}
a <-c(1,1)
b <- c(-1,1)
sum(a*b)

```

#### Length (norm) of a vector

In mathematics, the inner product of a vector with itself is called the *length* or *norm* of the vector:

```{r}
sqrt(sum(a*a))

```

The *length* of the vector $\mathbf{a}$ is obviously $\sqrt{2}$ .

*This terminology can be a bit confusing, because within R* `length()` *refers to the number of elements of a vector. Try to be as clear with your language as possible.*

*Tip: Remember, if you lose track of what variables you have defined, you can find them in the Global environment (tab Environment in the top right panel in RStudio, and accessible using* `ls()`*).*

## Matrices in R

A basic introduction to defining and manipulating matrices is given here. Defining a matrix starts by defining a vector containing all values. Then to define a matrix, the vector is split into column vectors by giving the number of columns:

```{r}
A <- matrix(c(1,2,3,4,6,5,7,8,9),ncol=3)
A
```

Here is another example:

```{r}
A2 <- matrix(c(1,2,3,4,6,5,7,8,9),nrow=2)
A2
```

You can see here an example of things leading to a warning message. R still returned a result. *Try to find out, what R is doing*.

### Accessing matrix elements

Here you could see that to access elements in a matrix, the values before the comma refer to the rows of the matrix and after the comma to the columns. The first column of a matrix can be assessed by:

```{r}
A[,1]
```

The second row by:

```{r}
A[2,]
```

And a sub-matrix by:

```{r}
A[2:3,2:3]
```

### Some useful matrices

The *identity matrix* of size 3 ( $\mathbf{I}_3$ ) is given by:

```{r}
diag(rep(1,3))
```

To fill a matrix of specified size with ones, do:

```{r}
B <- matrix(1, nrow = 4, ncol = 5)
B
```

*This time is as good as any to say that R knows both named and positional arguments. This means, you could have just as well defined the matrix like so:*

```{r}
B <- matrix(1, 4, 5)
B
```

*Look at the help function for* `matrix()` *to figure out the order of these arguments and other useful arguments for matrix definition*.

*Tip: Sometimes you would like to remove all your data and start over. You do this with the broom in the global environment, or by* `rm(list=ls())` .

### Linear Algebra in R

In R it is possible to manipulate vectors and matrices. You have already seen how to add two vectors and the multiplication of a vector and a constant.

You've also already seen the length of a vector, which you can calculate using the `norm()` function. E.g. according to Pythagoras, the length of the vector $\left[ 3\text{ }4\right]$ equals $\sqrt{3^2 + 4^2} = 5$ .

```{r}
v34 <- c(3,4)
lv <- norm(v34, "2")
lv
```

*Check the help for* `norm()` *to see, why you have to give the* `"2"` *argument*.

#### Vector multiplication

We start with multiplication of two vectors. *Remember, normally in linear algebra vectors are defined as columns.*

```{r}
v <- c(1,2,3)
b <- c(2,4,6)
v
b
```

If we do

```{r}
v*b
```

this is the element-wise multiplication.

Remember from above, there is also the *inner product* of two vectors: You ask R to perform *matrix multiplication* given by `%*%`. This multiplication of vectors (and also matrices) must follow strict rules: *The number of columns of the thing on the left must be equal to the number of rows of the thing on the right* of the multiplication symbol:

```{r}
t(v) %*% b
```

This *inner product* of two vectors is also called the *dot product*. Geometrically, the dot product is defined as $\mathbf{a}\cdot\mathbf{b} = ||\mathbf{a}||\text{ }||\mathbf{b}||\text{ }cos \theta$ (read: the length of the first vector times the length of the second vector times the cosine of the angle between the two vectors). A very nice YouTube tutorial can be found [here](https://www.youtube.com/watch?v=QPkKWGq_V0U).

If two vectors are *orthogonal* (their angle = 90°), the inner product of these two vector equals 0 (as cosine of 90° = 0). If the lengths of orthogonal vectors equal 1, then we call them *orthonormal*.

A matrix consisting of only orthonormal vectors is called an *orthogonal matrix.*

You get the *outer product* of two vectors by:

```{r}
C <- v %*% t(b)
C
```

You see that the outer product $\mathbf{C}$ is a matrix, while the inner product of two vectors was a scalar. The *rank* of the outer product matrix is 1, as it is the result of the multiplication of one pair of vectors (matrices with a single column -- or row). Thus, the rank of a matrix represents *with how many pairs of vectors it can be reproduced.*

#### Vectorized functions

Remember the element-wise multiplication from above? The same works with division:

```{r}
v/b
```

If you pass a vector to many predefined math functions in R, the function will return a vector of the same size, and each entry is found by performing the specified operation element-wise on the corresponding entry of the original vector.

```{r}
sin(v)
```

```{r}
log(v)
```

Now complex operations can be defined that can be done quickly and easily. In the following example a very large vector is defined and can be easily manipulated.

```{r}
x <- seq(0,pi,by=0.01)
y <- sin(x)*x/(2+cos(x))
plot(x,y)

```

#### Matrix multiplication

Multiplications of matrices with vectors or matrices with other matrices follow the same rules as vector multiplications. *The number of columns of the first matrix must be the same as the number of rows of the second*. Actually, multiplication of a matrix with a vector is just the dot product of each row of that matrix times the vector. See?

```{r}
A <- matrix(c(1,3,6,2,4,7,3,5,8),ncol = 3)
A
```

```{r}
v <- array(c(3,0,-1))
v
```

*Remember that this vector is a column vector.*

```{r}
A %*% v
```

Thus, the first element is the dot product of `A[1,]` and `v`. In this case this is 0 as $[1\text{  }2\text{  }3]$ and $[3\text{  }0\text{ }{-1}]$ are orthogonal. Multiplication of two matrices is just the repetition of a matrix and a vector for each vector in the second matrix:

```{r}
v2 <- cbind(v,2*v) # two vectors are combined into a matrix by sticking the columns together
v2
```

```{r}
A %*% v2
```

#### Interpretation of a matrix multiplication

A matrix $\mathbf{X}$ (of size $I\times J$) can transform a vector from $R^J$ into $R^I$, as $\mathbf{y} = \mathbf{Xb}$ . As $\mathbf{X}$ has $J$ columns, it can only be post-multiplied by vectors of size $J\times 1$ (such as $\mathbf{b}$) and the result is always a vector such as $\mathbf{y}$ (of size $I$). Thus, $\mathbf{X}$ transforms a vector of size $J$ (which is a point in the $R^J$ space) into a vector of size $I$ (which is a point in the $R^I$ space).

#### Inverse of a matrix

Now would it be possible to do the opposite? Can we transform $\mathbf{y}$ back to $\mathbf{b}$? For that, we would need a matrix $\mathbf{Z}$ of size $J\times I$ such that $\mathbf{b} = \mathbf{Zy}$.

This leads to $\mathbf{b} = \mathbf{Z} \mathbf{Xb}$. This means that $\mathbf{ZX}$ should be the identity matrix. In such a case $\mathbf{Z}$ is called the *inverse of* $\mathbf{X}$ (indicated as $\mathbf{X}^{-1}$). Only *square matrices* (with equal number of rows as columns ($I=J$)) can have an inverse.

*Intermezzo: If* $\mathbf{X}$ *is* $2\times 3$, *then* $\mathbf{b}$ *is a vector of size 3 and* $\mathbf{y}$ *is of size 2. Now it is* *impossible to make any vector* $\mathbf{b}$ *(of size 3) from a vector* $\mathbf{y}$ *of size 2. All solutions* $\mathbf{b}$ *would be in a 2d plane in* $R^3$.

Only, if $\mathbf{X}$ is square, it could have an inverse $\mathbf{X}^{-1}$, such that $\mathbf{X}^{-1} \mathbf{X} = \mathbf{I}$ (the identity matrix).

*If this sounds weird, think of it like this: t*he inverse of a matrix is similar of the inverse of a scalar. If you multiply the inverse of 5 (= $\frac{1}{5}$) times 5, the answer is 1.

#### Rank of a matrix

We already discussed the rank of a matrix earlier. The rank of a matrix relates to solving a homogeneous system of linear equations, or $\mathbf{Xb} = 0$, which is discussed in the Math syllabus. The rank is the number of vector pairs that reproduce the matrix. The maximum rank of a matrix is the minimum of its size. Thus, a $3\times 2$ matrix $\mathbf{X}$ has maximum rank 2 (but could also be rank 1). If two matrices are multiplied, the rank of the product is at maximum the lowest rank of the two matrices. Thus, the rank of $\mathbf{XX}^T$ equals the rank of $\mathbf{X}^T\mathbf{X}$ and equals the rank of $\mathbf{X}$ and it is maximally 2. For a matrix $\mathbf{X}$ of size $3\times 3$, it only has an inverse, if the rank of $\mathbf{X}$ is 3. This is called *full rank*.

Matrix $\mathbf{A}$ , also of size $3\times 3$, which was defined above, is not full, rank; thus there is no inverse. In R:

```{r, error=TRUE}
solve(A)

```

the `solve()` function returns an error.

*If it feels a bit strange to use the* `solve()` *function here, what do you solve for?* *check out* `?solve` *and see the explanation for the* `b` *argument* (*if you don't specify what to solve it puts the Identity matrix of the same size as the first argument as the second argument. So actually you solve* `A %*% ???? = I`).

You could have also checked whether $\mathbf{A}$ is of full rank by:

```{r}
y <- qr(A)

y$rank
```

You see that the rank is only 2 and not 3, thus there is dependency between the columns or rows.

*To explain,* `qr()` *performs a so-called QR decomposition (which we will not use any further here) and one of the results is the rank.*

#### Singular value decomposition

As discussed above, a matrix can be reproduced by a certain number of vector pairs that depend on the rank of that matrix. The *singular value decomposition* aims to find a minimum set of vectors that can reproduce the matrix as well as possible:

If $\mathbf{X}$ is a $I\times J$ matrix, then it is decomposed into sets of vectors of size $I$ (called $\mathbf{u}$) and of size $J$ (called $\mathbf{v}$), and how much of the variation in the data can be described with this pair (called $\mathbf{d}$). $\mathbf{u}$ and $\mathbf{v}$ are called *singular vectors* and $\mathbf{d}$ is the *singular value*. There are many additional properties of the SVD, which we will discuss later in the course. A nice example is the approximation of an image using SVD.:

```{r, error=TRUE}
install.packages("tiff") #this is how additional packages are installed
library("tiff") #this is how we load them
```

*Note: There are many R packages containing specialized functions. Packages can also be installed via the "Packages" tab in RStudio's lower right panel, searching/selecting a package, and clicking "Install". Making the package available in the workspace can be done in the console by the* `library()` *command or by ticking the appropriate box in front of the package.*

```{r}
cam <- readTIFF('cameraman.tif', 
                native = FALSE, 
                all = FALSE, 
                convert = FALSE, 
                info = FALSE, 
                indexed = FALSE, 
                as.is = FALSE)
```

*If the above chunk does not work, you need to move/copy the file* `cameraman.tif` *into your current working directory (accessible by* `getwd()`). *You may also want to change your working directory (using* `setwd()`*).*

Decompose the image using the svd function:

```{r}
USV <- svd(cam)
```

Re-construct the image from `N` singular vectors/values:

```{r}
N <- 5
cam2 <- USV$u[,1:N] %*% diag(USV$d[1:N]) %*% t(USV$v[,1:N])

# a tif file should have value between 0 and 1 so I need to do a correction
cam2[cam2<0] <-0
cam2[cam2>1] <-1

writeTIFF(cam2,"cam2.tif") 
# by default the file cam2.tif is stored in your working directory.
```

Now inspect this by a viewer. Try other values of N and rerun the code to see how well you can approximate the image.

## Some more R data types, functions, and simple plotting

#### Data frames and lists

Besides vectors and matrices, R also has the option to store values in a *data frame* and a *list* structure.

A data frame is used for storing data tables. It is a list of vectors of equal length. For example, the following variable `df` is a data frame containing three vectors `number`, `string`, `binary`.

```{r}
number <- c(2, 3, 5) 
string <- c("aa", "bb", "cc") 
binary <- c(TRUE, FALSE, TRUE)
quant = c(1,2,3,4)
df <- data.frame(number, string, binary)       
df
```

Specific elements of the data frame `df` can be called by (column) number or name:

```{r}
df[,1]
df$number
```

A list structure is more flexible than a data frame as it does not require that all variables have equal length.

```{r}
l = list(first=number,second=string,b=binary,last=quant)
```

List elements can also be called by name (or number):

```{r}
l$last
l[["last"]]
l[[4]]
```

#### Making plots

R can also let you graph results. The basic command is `plot()`:

```{r}
x <- 0:10
y <- sqrt(x)
plot(x,y)
plot(x,y, type = "l")
```

*Check out* `?plot` *and also* `?par`. *In this course, understanding your data analysis is more important than perfect plots - you will engage with that in later courses. However, basic R plotting is extremely powerful, if you know a few tricks.*

You can often program R to extract additional things to plot. We will use the example of moving average smoothing (*do you understand what that means?*):

```{r}
p <- seq(0,pi, by = 0.05) # values between 0 and pi
lp <- length(p)
Y <- sin(p) + rnorm(lp)*0.03 # simulated noisy measurements

M <- 5 #window length
dM <- floor(M/2)
startpos <- ceiling(M/2)
endpos <- lp-dM
S <- array(0,lp)
for(i in startpos:endpos){ 
  S[i] <- mean(Y[(i-dM):(i+dM)])
}
```

Now, we have a measured (noisy) vector and the smoothed solution. To compare the two, the measured values are plotted with circles, while the smoothed version is plotted as a line. We will use the `par()` command to make more than one plot in the same figure. The `mfrow(2,1)` makes 2 panels on top of each other, filling them from top to bottom.

```{r, fig.height=5,fig.width=3}
par(mfrow=c(2,1),mar=c(3,3,2,0.5),mgp=c(2,0.5,0))
plot(p,Y, xlab = 'angle',ylab='intensity', main = 'moving average')
lines(p[startpos:endpos],S[startpos:endpos],las=1)
plot(p[startpos:endpos],Y[startpos:endpos]-S[startpos:endpos],xlab = 'angle', ylab= 'deviation',main='Residuals',col="red",las=1)
```

#### Saving and re-loading data

As you work through a session you generate all sort of data. The next question is how do you save your work? You can save it as an R-project with the "Save" pictogram in the right upper "Environment" screen. Under the "File" menu you can open a project.

Within R you can read and write files in different formats. To give an example:

```{r, error=TRUE}
data <- read.csv(file='filename.csv')
```

reads a csv file and puts the information into an object called `data`. In a similar way you can write csv file with the command `write.csv()`.

*Tip: if you want to save R objects that you only want to re-open in R, use the* `saveRDS()` *and* `readRDS()` *functions, which are way faster and consume less space than reading from text files.*

#### Scripts

A script file is a set of R commands stored in an R file. For example, if you have a publication the next script can produce your figure:

```{r}
x <- 0:10
y <- sqrt(x)
plot(x,y)
plot(x,y, type = "l")
```

Try this out by copying and pasting this snippet into a new .R file. Save the file and run (*source*) it by selecting "Source" in the "Code" menu.

#### Functions

Commonly, your scripts will contain your custom functions. Next to the many functions pre-defined in R and its many packages, you can also write your own function and save them. An example:

```{r}
Inner <- function(x,y){
  sum(x*y)
}
```

*Once you've run this definition, it is available in the workspace of this session.*

```{r}
Inner(c(0,1),c(1,0))
```

*You can also save it in a new script file.*

*Remember the* `norm()` *function that had the* `type` *argument that we always need to remember to set in order to get the 2-norm? Can you write your own function* `my_norm()` *that returns the norm of a vector as explained before?*

## EXERCISE

Make two column vectors of size (6 x 1); c1 = (1, 2, 3, 4, 5, 6) and c2 = (3, 2, 1, 6, 5, 4). Make two vectors of size (10 x 1): r1 = (1, 2, 3, 4, 5, 0, 0, 0, 0, 0) and r2 = (5, 4, 3, 2, 1, 1, 2, 3, 4, 5)

a)  Calculate the lengths of vectors c1 and c2.
b)  Calculate the inner product of c1 and c2.
c)  Mean-center c1 and c2.
d)  Calculate the outer product of the mean-centered c1 and r1. Do the same for the mean-centered c2 and r2. Add the two outer products together and call this X.
e)  Apply a singular value decomposition to X.
f)  Make scatter plots of the first two left singular vectors (u) and also of the first two right singular vectors (v). Put appropriate labels on the axes of the scatter plots.
g)  Look at the singular values of X. What can you about the rank of X?

## Some useful programming tools

### Control structures

#### if-(else)-statement

```{r}
w <- 3
if( w < 5 ) {
 d <- 2
 }else{
 d <- 10
}	
```

#### for-loop

```{r}
s <- rep(0,20)
h <- 1:9
for(i in 2:10){
  s[i] <- h[i-1] * 10
}
```

#### sapply and lapply

Because for-loops are very inefficient in R, a common alternative for many situations are offered by `lapply()` and `sapply()`, which apply a function to a vector (or list) and return a list or simplified list (vector or matrix), respectively, with the result for each element.

```{r}
s <- rep(0,20)
h <- 1:9
s[2:10] <- sapply(2:10, 
                  function(i){
                    h[i-1] * 10
                  })
```

### Centering and Scaling of a data matrix

In multivariate analysis of data one often mean centers the data. This means that the mean of each column (variable) is removed from each value in that column. This can be applied as follows. We will use the cam matrix from the image above

```{r}
# Here you calculate the mean of
mean_cam = colMeans(cam)  # The colMeans function calculates the mean of each column in X
cam_meancentered = sweep(cam,2,mean_cam,FUN = "-") # 
```

The sweep function is used. It's input values are: X, the matrix on which its function is applied; 2, the dimension on which it is applied, here 1 means apply it on the row dimension, 2 means apply it on the column dimension; meanX, the vector with means; FUN = "-", means that the subtract function is used.

Similarly each variable of the data matrix is also often scaled using the standard deviation of that variable. Thus each element of a column is divided by the standard deviation of that column. This can be applied as follows:

```{r}
std_cam = apply(cam_meancentered,2,FUN = sd) # Unfortunately there is no colStd function, you have to use the apply function
cam_standardized = sweep(cam_meancentered,2,std_cam,FUN = "/") 
```

The combined mean centering and scaling by its standard deviation is called standardization or autoscaling of X.

## EXERCISE 2

-   Create a data matrix of 5 rows and 4 columns and fill it with some numbers.
-   Calculate the mean and standard deviation of each column of that matrix.
-   Autoscale the data matrix
-   Recalculate the original data matrix using the autoscaled data, and the means and standard deviations of each column.

### Useful functions for data creation

*Remember, you can access a description of how to use all of these functions in the Help tab, or using* `help(functionName)` *or* `?functionName`.

#### `read.table()`

read a table from file. *Arguments:*

-   `header=TRUE`: read first line as titles of the columns

-   `sep=","`: numbers are separated by commas

-   `skip=n`: don't read the first n lines.

#### `write.table()`

write a table to file.

#### `c()`

concatenate numbers (or other objects) to create a vector

#### `vector()`

create a vector. *Arguments*:

-   `mode="numeric"`: data type of the elements

-   `length=""`: number of elements

#### `matrix()`

create a matrix. *Arguments*:

`nrow` and/or `ncol`: number of rows/columns

#### `data.frame()`

create a data frame

#### `list()`

create a list

#### `rbind()` and `cbind()`

combine vectors into a matrix by row or column

### Information on variables

#### `length()`

length of a vector (or a list)

#### `ncol()` or `nrow()`

number of columns or rows in a matrix (or data frame)

#### `class()`

class of a variable

#### `names()`

names of objects in a list (or a vector)

#### `print()`

show variable or character string on the screen (used in scripts or for-loops)

#### `return()`

return variable from a function. Shows variable on the screen, unless it's assigned to another variable.

#### `is.na()`

test if variable is `NA`

#### `as.numeric()` and `as.character()`

change class to number or character string

### Extracting data

```{r eval = FALSE}
x[n] # the nth element of a vector
x[m:n] # the mth to nth element
x[c(k,m,n)] # specific elements
x[x>m & x<n] # elements between m and n
x$n # element of list or data frame named n
x[["n"]] # idem
x[i,j] # element at ith row and jth column of a matric or data frame
x[i,] # row i in a matrix
x[,j] # column j in a matrix
```

### Statistics

#### `sum()`

sum of a vector (or matrix)

#### `mean()`

mean of a vector

#### `sd()`

standard deviation of a vector

#### `max()` or `min()`

largest or smallest element

#### `rowSums()` - or `rowMeans()`, `colSums()` and `colMeans()`

sums (or means) of all numbers in each row (or column) of a matrix. The result is a vector.

#### `quantile(x,c(0.1,0.5))`

sample the 0.1 and 0.5th quantiles of vector x

### Plotting

#### `plot(x)`

plot x (y-axis) versus index number (x-axis)

#### `plot(x,y)`

plot y (y-axis) versus x (x-axis)

#### `image(x,y,z)`

plot z (color scale) versus x (x-axis) and y (y-axis)

#### `lines()` or `points()`

add lines or points to a previous plot. *Check out the help for* `points()` *to learn which symbols R knows about.*

#### `hist()`

plot histogram of the numbers in a vector

#### `barplot()`

bar plot of vector or matrix

#### `contour(x,y,z)`

contour plot

#### `abline()`

draw line (segment). *Arguments*:

-   `a,b` for intercept a and slope b; or

<!-- -->

-   `h=y` for horizontal line at y; or

-   `v=x` for vertical line at x.

#### `curve()`

Plot a function. Needs to have an x in the expression, for example `curve(x^2)` . Use argument `add=T` to add to current plot.

#### `legend()`

add legend with given symbols, line type and colours (`pch`, `lty` and `col`) and text (legend) at a defined location (e.g., `x="topright"`)

#### `axis()`

add axis. *Argument* `side`:

1: bottom, 2: left, 3: top, 4: right

#### `mtext()`

add text on axis. *Arguments*:

-   `text` character-type axis label

-   `side`

#### `grid()`

add grid

#### `par()`

plotting parameters *Check out the help for this one to learn how to adapt nearly everything in simple plots.* *Arguments* to be specified before plotting:

-   `mfrow=c(1,3))` : number of figures per page (1 row, 3 columns)

-   `new=TRUE` : draw plot over previous plot.

*Plotting parameters that can be added as arguments* to `plot()`, `lines()`, `image()`, etc. For help see `?par`.

-   `type` : `type="l"` means lines,`type="p"` means points, etc.

-   `col` : color, e.g. `"blue"`, `"red"`, etc

-   `lty`: line type - `1` means solid, `2` is dashed, etc.

    -   `pch` : point type -- `1` is an open circle, `2` an open triangle, etc. (see`?points`)

-   `main` : title - character string

-   `xlab` and `ylab`: axis labels -- character string

-   xlim and `ylim`: range of axes -- e.g. `c(1,10)`

-   `log`: logarithmic axis -- `"x"`, `"y"`or `"xy"`

## Further reading

#### This tutorial is based on

<https://cran.r-project.org/doc/contrib/Torfs+Brauer-Short-R-Intro.pdf>

Here are a few more open access resources for R:

#### Shorter Reads

<https://www.rstudio.com/resources/cheatsheets/>

<https://www.r-bloggers.com/2015/12/how-to-learn-r-2/#h.nl3rvdr3ycsc>

The Undergraduate Guide to R - A beginner's introduction to the R programming language (Trevor Martin):

<https://www.biostat.jhsph.edu/~ajaffe/docs/undergradguidetoR.pdf>

Little Book of R for Biomedical Statistics Avril Coghlan

<https://a-little-book-of-r-for-biomedical-statistics.readthedocs.io/en/latest/>

#### R books

Modern Statistics for Modern Biology (Susan Holmes & Wolfgang Huber):

<https://web.stanford.edu/class/bios221/book/index.html>

Introduction to Data Science (Rafael A. Irizarry):

<https://leanpub.com/datasciencebook>

R for Data Science (Hadley Wickham & Garrett Grolemund):

<https://r4ds.had.co.nz/index.html>

R Programming for Data Science (Roger D. Peng):

<https://leanpub.com/rprogramming>

Data Analysis for the Life Sciences (Rafael A. Irizarry & Michael I. Love):

<https://leanpub.com/dataanalysisforthelifesciences>

Exploratory Data Analysis with R (Roger D. Peng):

<https://leanpub.com/exdata>

Applied Statistics with R (David Dalpiaz):

<https://daviddalpiaz.github.io/appliedstats/applied_statistics.pdf>

<http://daviddalpiaz.github.io/appliedstats/>

R Cookbook (James (JD) Long & Paul Teetor):

<https://rc2e.com/>

Advanced R (Hadley Wickham):

<https://adv-r.hadley.nz/>

#### Linear Algebra

Linear Algebra for Data Science with examples in R (Shaina Race Bennett):

<https://shainarace.github.io/LinearAlgebra/>

Excerpts: Introduction to Linear Algebra (Gilbert Strang):

<https://math.mit.edu/~gs/linearalgebra/>
